---
title: "A Review of A Multivariate Approach to the Study of Orichalcum ingots from the Underwater Gela's Archaeological Site"
format: 
  html:
    css: style.css
  pdf:
    include-in-header:
      text: |
        \usepackage{setspace}
        \doublespacing
editor: visual
bibliography: ../Common/CITATIONS.bib
csl: ../Common/apa.csl
nocite: |
  @*
---

```{r packages and set up, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(tabulapdf)
library(dplyr)
library(tidyr)
library(pdftools)
library(ggcorrplot)
library(mvtnorm)
library(GGally)
library(here)
library(tinytex)
library(httr)
library(MVN)
library(ggplot2)
library(cluster)
library(factoextra)
library(hexbin)
library(matrixStats)
library(fclust)
library(knitr)
library(tidyverse)
library(skimr)
library(bibtex)

my_col <- c("#648FFF", "#785EF0", "#DC267F", "#FE6100","#FFB000","#F0E442")

##  Reading the pdf: - pdftools package code:

#Load paper by caponnetti et. al from repository and store locally temporarily

url <- "https://raw.githubusercontent.com/luke8471/STAT394-GROUP-3/02d02760e329f00dd8cc8cc8543327ffeccd00a6/Text/Articles/caponetti%20et%20al.pdf"

tmp <- tempfile(fileext = ".pdf")

pdf <- httr::GET(url, write_disk(tmp, overwrite = TRUE))

ingots <- pdftools::pdf_text(tmp)

```

*Gaia Smith, Alice Thompson and Luke Harkness*

## 1. Abstract

General Background

A shipwreck off the coast of Gela, Italy has been the source of a historical find of 40 orichalcum ingots. A scarce, fabled metal that has the modern equivalent of brass.

sPECIFIC BACKGROUND

Knowledge gap

Large groupings and finds from one source are very hard to come by: lots of data in one place.

Here we show...

After performing a trilogy of cluster analyses, we show that were the clusters accurate were we able to determine a good fit?

Our goal is to create and present a comprehensive, comprehensible, reproducible, and replicable multivariate analysis of the data obtained from the chemometric elemental analysis of 38 orichalcum ingots that were discovered in a 6th Century BC shipwreck off the coast Gela, Italy in 2015.

Results

Implications

They had different origins. Can possibly be linked to geological locations by ore analysis, collected from different regions, etc.

Why did they wait to do further testing that would provide more information on the composition and provide more accurate data?

The groupings from the @Caponetti2017-fd study appears to have been manipulated without compensation of the . By taking the log at any point in this anylysis without standardisation would have resulted in multiple zero-values and artificially inflating

Is the cluster analysis consistent with different distances and linkages?

This is the largest sample of its kind however it is still not a large sample.

For the analysis of our report we will be considering all the main components and trace element concentrations excluding the Cu/Zn ratio as...

Why do we care about the problem and the results?

Because it could lead to further impressive archaeological finds

What is the scope of the work and what problem is this paper trying to solve?

Determine whether or not any of the ingots could be grouped. The scope is mapping orichalcum origin.

What was done to solve this problem?

More destructive testing to determine trace elements.

What implications does the answer imply?

one of our hypotheses questions is whether our data follows a multivariate normal distribution. probably some hypotheses around the clusterings.

------------------------------------------------------------------------

## 2. Introduction

**2.1 Significance of this archaeological find and Gela**

The historic Gela, Italy has been a treasure trove of archaeological finds for generations. This ancient city was strategically positioned at the crux of major maritime trade routes, which made it a populous and successful trading port [@Angelucci2026]. Consequently, ships looking to trade goods like pottery, metals, and grain have traveled to and from Gela for centuries.

A shipwreck off the coast continues to provide significant archaeological finds to this day. This discovery dates back to 1988 [@nuttersworldGelaWreck], where it was first stumbled upon by divers near Bulala beach. We are exploring one of the most significant finds from this wreck; A veritable hoard of the fabled metal, Orichalcum.

The discovery of 40 orichalcum ingots in 2017 was a monumental one. There had never been a discovery of this quantity in one place anywhere in the world. There was also evidence of consistent casting processes, with overall similarities suggesting a rudimentary mold production. This was likely a hole dug in the ground and molten alloy was poured in the indentation to solidify. The morphology of the found ingots suggested cast-diverse technologies.

**2.2 What is orichalcum?**

Orichalcum is defined as a copper alloy with zinc content as the principal components. A widely varying quantity of lead and tin among other trace elements are expected. There is no consistent proportion to the amount of copper and zinc found, but the overall composition and relative amounts of principal elemental components can be used to deduce the rough date and origin of a sample. The modern day equivalent is brass.

**2.3 Summary of Collection Techniques**

In the initial analysis of the orichalcum ingots, X-Ray Fluorescence (XRF) spectroscopy via a portable spectrometer [@Caponetti2017-sc] was used. The subject paper includes that data and further investigates the composition of each ingot using more invasive and refined elemental analysis.

To obtain material for further testing, a micro-well was drilled into each ingot and the extracted alloy was taken to determine trace elements using Inductively coupled plasma optical emission spectroscopy (ICP-OES) and Inductively coupled plasma optical emission spectroscopy (ICP-MS) of each of the 38 orichalcum ingots. (FLESH OUT AND EXPLAIN MORE, TOO VAGUE NOT EXPLAINING WHY)

**2.4 Scope of This Multivariate Analysis**

This multivariate analysis includes manipulations of the data collected from the ICP-OES and ICP-MS chemometric analysis, it does not include the data about physical characteristics of the ingots or any correlation analysis of those variables.

We will be excluding the Cu/Zn ratio variable as we chose to analyse the linkage of all elements between each other, not between Cu/Zn ratios and other elements

------------------------------------------------------------------------

## 3. EDA

**3.1 Analysing raw data**

We performed a simple summary of the data in table 1 and creating a box plot to visualize the data. Something to note is that the elements Cu, Zn, Pb, and Fe are all in percentages, where as the concentrations of the other elements are in PPM (parts per a million). This immediately indicated that scaling was needed.

```{r extracting the table, echo=FALSE}
#| fig-width: 10
#| fig-height: 6

##  Extracting the table from the pdf: - tabulapdf package codes:
ingots2 <- suppressMessages(extract_tables(url, pages=3))

#lets store it in a table:

table1 <- suppressMessages(ingots2[[1]])

#when I first did this this you couldn't view Ni, Ag and Sb separately they all went under concentration
##  Cleaning the table: - DATA202 codes
table1 <- suppressMessages(table1 %>%
                             separate(
                               'Concentration in PPM',
                               into = c("Ni", "Ag", "Sb"),
                               sep = " "
                             ))

#this should now have all the columns separate :)

#making sure all the columns are named correctly:
suppressMessages(names(table1) <- c(
  "Sample", "Cu", "Zn", "Pb", "Fe",
  "Ni", "Ag", "Sb", "As", "Co", "Cd", "Sn",
  "Te", "Bi", "Mn", "Li", "Al", "V", "Cr",
  "Rb", "Sr", "Ba", "Cu_Zn"
))

#as R was having a hard time using the table above we created a Data Frame of the Concentrations:

Cu <- as.numeric(table1$Cu[-1])
Zn <- as.numeric(table1$Zn[-1])
Pb <- as.numeric(table1$Pb[-1])
Fe <- as.numeric(table1$Fe[-1])
Ni <- as.numeric(table1$Ni[-1])
Ag <- as.numeric(table1$Ag[-1])
Sb <- as.numeric(table1$Sb[-1])
As <- as.numeric(table1$As[-1])
Co <- as.numeric(table1$Co[-1])
Cd <- as.numeric(table1$Cd[-1])
Sn <- as.numeric(table1$Sn[-1])
Te <- as.numeric(table1$Te[-1])
Bi <- as.numeric(table1$Bi[-1])
Mn <- as.numeric(table1$Mn[-1])
Li <- as.numeric(table1$Li[-1])
Al <- as.numeric(table1$Al[-1])
V <- as.numeric(table1$V[-1])
Cr <- as.numeric(table1$Cr[-1])
Rb <- as.numeric(table1$Rb[-1])
Sr <- as.numeric(table1$Sr[-1])
Ba <- as.numeric(table1$Ba[-1])
Cu_Zn <- as.numeric(table1$Cu_Zn[-1])

Ingots_Frame <- data.frame (Sample = c("S1","S2","S3","S4","S5","S6","S7","S8","S9","S10","S11","S12","S13","S14","S15","S16","S17","S18","S19","S20","S21","S22","S23","S24","S25","S26","S27","S28","S29","S30","S31","S32","S33","S34","S35","S36","S37","S38"),Cu, Zn, Pb, Fe, Ni, Ag, Sb, As, Co, Cd, Sn, Te, Bi, Mn, Li, Al, V, Cr, Rb, Sr, Ba, Cu_Zn)

knitr::kable(Ingots_Frame,
             caption = "Figure 1: Table 1 The main components and trace elements from Caponetti et al study",
             format = "pipe")

# Removing Cu/Zn from the table
Ingots_F_no <- Ingots_Frame[,1:22]
```

looking at the raw data we also noticed some interesting outliers. Particularly, sample 13 for Sb (antimony), which is \mathcal{549PPM}, magnitudes larger than the other samples. (INSERT MORE OBSERVATIONS FROM MAIN TABLE HERE) a large range of concentrates (ranging from a minimum of 0.06PPM for Rb and a max of 683PPM for Ni)

```{r basic summary, echo=FALSE}
#| fig-width: 10
#| fig-height: 6

#Summary of the Data Frame:
invisible(summary(Ingots_F_no))

#code to make the summary into a table
skim(Ingots_F_no[,-1]) |>
  select(-n_missing, -complete_rate)
  

#box plot of the initial data: (no transformations) using ggplot:
#start by turing the data into long format:
Ingots_long <- Ingots_F_no[,-1] |>
  pivot_longer(
    cols = everything(),
    names_to = "Elements",
    values_to = "Concentrations")



ggplot(Ingots_long, aes(x= Elements, y= Concentrations)) +
  geom_boxplot(notch = FALSE, fill ="#40B0A6") +
  theme_minimal()+
  labs(title = "Boxplot the table 1")

```

Ni (nickel) had the widest interquartile range, indicating the greatest variation among the samples, while Sb (antimony) shows the largest total range, likely driven by sample 13's recorded Sb being magnitudes larger than any of the other samples.

Due to the fact the data was recorded at different scales (percentage and PPM) and obvious outliers, log transformation and scaling were the next logical steps.

***3.1.1 Inconsistencies in Percentages***

The percentages of elements Cu, Zn, Pb, and Fe do not add up to 100% for each sample, as shown in the table below.

```{r suprising, echo=FALSE}
#| fig-width: 10
#| fig-height: 6

main_com <- Ingots_Frame[c(1:5)]

#adding the sum of the main components:
main_com$percentage_total <- rowSums(main_com[c(2:5)])

#table of the main components:
knitr::kable(main_com,
             caption = "Deeper dive into the main components",
             format = "pipe")

#summary of the main components:
skim(main_com[,6])|>
  dplyr::select(-n_missing, -complete_rate)

```

These samples have a mean of \\numcal{104.4%} with a range \\numcal{93.17%} to \\numcal{113.71%}. Where 33 of the 38 samples exceed \\numcal{100%}. The most significant deviations are samples 8 and 38 (COMMENT ON THIS). @Caponetti2017-fd attribute this to be due to errors generated from the low quantity of available material for sampling for each observation.

**3.2 QQ-plots & Henze-Zirkler Test For Multivariate Normality**

We first compared the QQ-normal plots of table 1 and compared it to the log transformed table 1 to assess if log transformation would be appropriate to manage outliers and standardisation.

```{r qqnormal, echo=FALSE}
#| fig-width: 10
#| fig-height: 6

#no transformations: elements in percentages:

qqnorm(as.matrix(Ingots_F_no[-1,2:5]), col="#648FFF", main = "Normal QQ-Plot of Table 1", sub =  "Using all elements that were recored in percentages (Cu, Zn, Pb and Fe)")
qqline(as.matrix(Ingots_F_no[-1,2:5]), col="red")

#no transformations: elements in PPM:

qqnorm(as.matrix(Ingots_F_no[-1,6:22]), col="#648FFF", main = "Normal QQ-Plot of table 1", sub = "Using all elements that were recored in PPM")
qqline(as.matrix(Ingots_F_no[-1,6:22]), col="red")

```

As this was before there was any scaling performed, we opted to display the elements measured in percentages and the elements measured in PPM separately to better display their individual influence.

Percentages: Shows significant departure from normality...

PPM: Shows significant departure from normality....

The QQ-plot of the raw data had some clear deviations in the tails (lower and mostly upper data points), few of the points actually follow the theoretical line (in red), this alone indicated that the normality assumption may not hold. Their are also some slight steps in the points as well as a large jump at (THIS IS ALL NEEDS TO BE REVISED)

```{r}
#| fig-width: 10
#| fig-height: 6

#log transform data

logIngots <- log(as.matrix(Ingots_F_no[-1,-1]))
#print(logIngots)

knitr::kable(logIngots,
             caption = "Log Transformed Data",
             format = "pipe")

#taking the log.

qqnorm(logIngots, col = "#785EF0", main = "Normal QQ- Plot of the log-transformed data (all elements)")
qqline(logIngots, col = "red")

```

The QQ-plot of the log transformed data roughly followed the theoretical line (in red), this is evidence that the normality assumption may hold. The tails are relatively long and flat, indicating that there may be some extreme outliers. There are several steps in the upper and lower tails, which alone is not necessarily a departure from normality, and may be explained by the zero values returned for some Sample's variables after the log transformation.

The observed zero-values for (PUT SAMPLES NUMBERS IN HERE) in (TABLE) in variables(HERE), which while not a large quantity of the samples, still provided some concern as \numcal{log(0)} is undefined and can influence non-normality in multivariate data. The negative values are not an issue because we will be using metrics which are calculated from coordinate distances for clustering, and can handle negative values.

We chose to combat the zero values by standardising the data set using \mathcal{log(x + 1)}, where \mathcal{x} is our data. This, coupled with a Henze-Zirkler test, established if there is preliminary evidence for multivariate normality.

```{r}
#| fig-width: 10
#| fig-height: 6

#log transform data

logIngots1 <- log(as.matrix(Ingots_F_no[-1,-1] + 1))

#print(logIngots)

knitr::kable(logIngots1,
             caption = "Log Transformed Data",
             format = "pipe")

#taking the log.

qqnorm(logIngots1, col = "#785EF0", main = "Normal QQ- Plot of the log-transformed data (all elements)")
qqline(logIngots1, col = "red")


HZoriginal <- mvn(data = logIngots1, mvn_test = "hz")
cat("Henze_Zirkler Test Results:\n")
print(HZoriginal$multivariate_normality)

```

As you can see in table 2, our values are now all non-zero.

The normal QQ-plot looks less normal after standardisation, with the length of the tails being accentuated and continuing presence of stepped data.

The Henze-Zirkler test stipulates

\begin{aligned}
\mathcal{H}_0 :& 
\text{ The data follow a multivariate normal distribution},
\\
\mathcal{H}_1 :& 
\text{ The data do not follow a multivariate normal distribution}.
\end{aligned}

This generated a \mathcal{p-value=0.069}, suggesting that there is insufficient evidence to reject that the data follow a multivariate normal distribution. However, the QQ-plot shows enough \<flat tails, step, steep, etc\> that further testing is warranted before determining if multivariate normality is likely.

Given the uncertainty of multivariate normality and the presence of outliers, further data manipulation by Medioid Absolute Deviation was determined to be the next best step towards multivariate analysis.

**3.3 MAD**

MAD was used because it is a robust, outlier-resistant, measure of variability. The log transformation somewhat fixed the skew and pulled the outliers in, but did not eliminate their mathematical influence.

```{r transformations-mad, echo=FALSE}
#| fig-width: 10
#| fig-height: 6

#Calculate medians and MAD for each column:
median <- apply(logIngots1[,-1], 2, median)
mad <- apply(logIngots1[,-1], 2, mad)


#print(median/mad)
#mad

med_mad_table <- data.frame(
  Median = median,
  Mad = mad,
  Median_over_Mad = median/mad
)

knitr::kable(med_mad_table,
             caption = "Medians / Mads",
             format = "pipe")

```

From {TABLE NUMBER} we observed some massive deviations, as expected.

Next the robust Z score was used to identify outliers that were candidates for removal. For our cutoff, we went with @Leys2013-cr recommendation, who argued that the MAD is the most robust scale measure under contamination and recommend a median plus or minus 2.5 time MAD rule for outlier detection.

```{r outliers, echo=FALSE}
#| fig-width: 10
#| fig-height: 6

#Compute the robust z-scores:
z <- scale(logIngots1[,-1], center = median, scale = mad)

#Flag observations where any variable exceeds the threshold
outliers <- apply(abs(z) > 2.5, 1, any)

#list(outliers)

# View the rows flagged as outliers
invisible(summary(logIngots1[outliers,]))

#The QQ plot of the mad transformed data:

qqnorm((z), col="#785EF0", main = "Normal QQ-Plot of 
       MAD-Transformed data")
qqline((z), col="red")

num_cols <- sapply(logIngots1, is.numeric)
z_scores <- as.data.frame(scale(logIngots1[, ]))

#summary(z_scores)

ol <- logIngots1[rowSums(abs(z_scores) > 4) > 0, ] #ALICE STUFF

print(ol) # ALICE STUFF

Ingots_clean <- logIngots1[rowSums(abs(z_scores) > 2.5) == 0, ]

#summary(Ingots_clean)

#New boxplot:

df <- as.data.frame(Ingots_clean)

Ingots_clean_long <- df[,-1] |>
  pivot_longer(
    cols = everything(),
    names_to = "Elements",
    values_to = "Concentrations")



ggplot(Ingots_clean_long, aes(x= Elements, y= Concentrations)) +
  geom_boxplot(notch = FALSE, fill ="#40B0A6") +
  theme_minimal()+
  labs(title = "Boxplot of MAD Transformation With Outliers Removed")


#still getting negative values (even with elevated magnitude cutoff) in here so 
#log transforming may not be appropriate.

logIngotsClean <- invisible(log(as.matrix(Ingots_clean[-1,-1])))
invisible(summary(logIngotsClean))

View(logIngotsClean)


#making the dataframe numeric so it can be manipulated.

num_vars <- sapply(Ingots_clean, is.numeric)

```

The QQ plot shows very skewed data which we would expected...

What is after we have "cleaned" up the outliers our \|z\| \> 3 rule removes samples 13, 25, and 38 which is exactly the sample ingots that Caponetti et al. set aside as their G4 group.

{should probably explain why cutting out 11 samples will be good for the rest of our analysis, something along the line of outliers, terrible samples and low sample quality}

***2.3.1 pairwise between CU, Zn, Pb, and Sn using the clean data***

```{r pairwise, echo=FALSE}
#| fig-width: 10
#| fig-height: 6

ggpairs(Ingots_clean, columns = c(2,3,4,12),
        lower = list(continuous = wrap("points", alpha = 0.6, size = 1.5, color = "#006CD1")),
        upper = list(continuous = wrap("cor", size = 4, color = "#005AB5")),
        diag = list(continuous = wrap("densityDiag", alpha = 0.5, color = "#40B0A6")))

```

Also done to reflect the traditional primary components that make up orichalcum, showing the relationship between these four elements.

**2.4 Multivariate normal distribution**

What tests did we use? Henze Zirkler

Why did we use those tests? #ALICE STUFF

```{R Henze Zirkler, echo=FALSE, warning = FALSE}
#| fig-width: 10
#| fig-height: 6

#HZ test on OG data, which also rejects null.

HZoriginal <- mvn(data = Ingots_F_no, mvn_test = "hz")
cat("Henze_Zirkler Test Results:\n")
print(HZoriginal$multivariate_normality)

#Henze-Zirkler test on cleaned data?
#Non-negative functional distance that measures between two distribution functions. If data
#is MVN, the test statistic HZ is approximately lognormally distributed. It calculates mean,
#variance, and smoothness parameter. Mean and variance are then log-normalised and p-value
#is estimated.
#@MISC{noauthor_undated-de,
#  howpublished = "\url{https://www.rdocumentation.org/packages/MVN/versions/5.8/topics/hzTest}",
#  note         = "Accessed: 2026-9-15"
#}

HZ <- mvn(data = Ingots_clean, mvn_test = "hz")

cat("Henze_Zirkler Test Results:\n")
print(HZ$multivariate_normality)
```

Due to our Henze Zirkler test for both the un-transformed data (p-value \<0.001) and the transformed data (p-value 0.046) we can safety reject the null hypotheses and conclude that the data regardless if it is transformed or not does not follow a multivariate normal distribution.

As our data isn't multivariate it isn't appropriate to perform a mahalanobis distance. {could potentially do need to research tho}

**2.5 Correlation plots before and after the transformation**

```{r correlations, echo=FALSE}
#| fig-width: 10
#| fig-height: 6

#Covariance table of the data:
invisible(var(Ingots_F_no[-1,-1]))


#Correlation table of the data:
invisible(cor(Ingots_F_no[-1,-1]))


#Correlation plot:
ggcorrplot(cor(Ingots_F_no[,-1]),
           hc.order = TRUE,
           type = "lower",
           lab = TRUE,
           lab_size = 2,
           ggtheme = theme_minimal(),
           colors = c("#5D3A9B", "white","#E66100"),
           tl.col = "gray30",
           title = "Correlation plot of the raw data (no transformations)")


#performed for clean data:
invisible(var(Ingots_clean[-1,-1]))

invisible(cor(Ingots_clean[-1,-1]))

ggcorrplot(cor(Ingots_clean[,-1]),
           hc.order = TRUE,
           type = "lower",
           lab = TRUE,
           lab_size = 1.5,
           ggtheme = theme_minimal(),
           colors = c("#5D3A9B", "white","#E66100"),
           tl.col = "gray30",
           title = "Correlation plot of the clean data exluding Cu/Zn")

```

Seeing a lot more correlation in the data with the massive outliers removed. Correlation plot some things to highlight; is that the strong positive association between V and Cr is also strongly positively association in the data with the outlying samples removes.. The transformed data also clearly shows a strong positive association for Bi and Ag which the un-transformed data doesn't show as significantly. the negative association between Zn and Te becomes more stronger in the transformed data.

The transformation didn't just strength some of the associations it also happened to reverse the sign for several of the relationships, Bi and Mn was one the raw data had a positive association but the transformed data turned it into a negative association similar with the relation ships between Sb and Sn as well as Rb and Te. It went the other direction (negative to positive) for As and Co. This is a positive sign that the robustness of the transformation, i.e removing some of the outliers demonstrates their influence on the raw data.

Given the uncertainty about multivariate normality, a linear data analysis was chosen over a primary component analysis.

## 4. LDA

**4.1 PAM**

We started by performing a PAM (Partitioning Around Medoids) analysis

```{r LDA, echo=FALSE}
#| fig-width: 10
#| fig-height: 6

ICN <- data.frame(sapply(Ingots_clean[,-1], as.numeric))

scaled <- scale(ICN)

#head(scaled, n = 3)

#need to find optimal silhouette width

sil_avg <- numeric(25)

for (k in 2:26) {
  PAM_fit <- pam(scaled, k)
  sil_avg[k - 1] <- PAM_fit$silinfo$avg.width
    }

LDA_K <- data.frame(k = 2:26, avg_silhouette = sil_avg)

knitr::kable(LDA_K,
             caption = "Figure 8: Table contain the k scaled values",
             format = "pipe")


pam <- pam(Ingots_clean, k = 5)

cluster <- pam$clustering

for (k in 2:26) {
  PAM_fit <- pam(scaled, k)
  sil_avg[k - 1] <- PAM_fit$silinfo$avg.width
    }

LDA_K1 <- data.frame(k = 2:26, avg_silhouette = sil_avg)

#summary(pam) ---- Not sure if we would like to include this or not

fviz_silhouette(pam)+
  scale_fill_manual( values = c("#648FFF", "#785EF0", "#DC267F", "#FE6100","#FFB000"))+
  scale_colour_manual( values = c("#648FFF", "#785EF0", "#DC267F", "#FE6100","#FFB000"))
  

```

The silhouette value/metric is a measurement of the quality of a clustering. between -1 and 1. The data reports the optimal silhouette width as 5, with 0.6496032, indicating some. There are several other clustering values with a reported similar quality.

Due to the ambiguous clustering, we opted to perform more cluster analyses to determine if this was a good fit.

MANHATTAN NOT EUCLEDIAN.

**3.3 Cluster Analysis**

H cluster using Manhattan Distance as a measure.

However we may not want to use wards method as .... {ALICE} ... hence Manhatten distance cluster may be a better cluster to consider as it more robust to outliers then Eucledian.

```{r manhatten cluster, echo=FALSE}
#| fig-width: 10
#| fig-height: 6

old_par <- par(no.readonly = TRUE) #stuff to help make the legend

dist <- dist(scaled, method = "manhattan")

HiMan <- hclust(dist, method = "average")

par(mar = c(6,5,4,6.5)) #stuff to help make the legend

plot(HiMan, main = "Manhatten Cluster Dendrogram")
rect.hclust(HiMan , 5, border = c("#648FFF", "#785EF0", "#DC267F", "#FE6100","#FFB000"))
legend("topright",
       inset = c(-0.10,0),
       legend = c("1", "2", "3", "4", "5"),
       fill = c("#648FFF", "#785EF0", "#DC267F", "#FE6100","#FFB000"), xpd = TRUE)

on.exit((par(old_par))) #stuff to help make the legend


```

Something interesting top note is that for both cluster options they both cluster sample's 15, 10, and 22 this could be because of ...

**3.4 Fuzzy Analysis**

As the third and final method of clustering, we performed a Fuzzy analysis.

```{r fuzzy, echo=FALSE}
#| fig-width: 10
#| fig-height: 6

old_par <- par(no.readonly = TRUE) #stuff to help make the legend

#k-medoids for Fuzzy

ICN <- Ingots_clean
scaled <- scale(ICN)

Fuzzy_k <- FKM.med(scaled, k = 5, m = 2)

#silhouettes

#print(Fuzzy_k$U)

#Table of Fuzzy_K$U:
knitr::kable(Fuzzy_k$U,
             caption = "Fuzzy K Silhouettes Clusterings",
             format = "pipe")



Fuzzy_Sil <- SIL.F(scaled, Fuzzy_k$U)

Fuzzy_Clust <- apply(Fuzzy_k$U, 1, which.max)

d <- dist(scaled)

sil_each <- silhouette(Fuzzy_Clust, d)

head(sil_each)

plot(sil_each,
     col = my_col[sort(unique(Fuzzy_Clust))],
     border = NA,
     main = "Silhouette Plot")


sil_df <- data.frame(
  obs = 1:nrow(sil_each), 
  width = sil_each[, "sil_width"],
  cluster = factor(sil_each[, "cluster"])
  )

ggplot(sil_df,
       aes(x = reorder(obs, width),
           y = width,
           fill = cluster)) + 
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = mean(sil_df$width),
             linetype = "dashed",
             colour = "red") +
  scale_fill_manual(values = c(
    "#648FFF", "#785EF0", "#DC267F", "#FE6100", "#FFB000")) +
  labs(title = "Silhouette Plot",
       x = "Observation",
       y = "Silhouette Width") +
  theme_minimal()

plot(sil_each[, "sil_width"],
     pch = 15,
     col = my_col[Fuzzy_Clust],
     xlab = "Observation",
     ylab = "Silhouette Width",
     main = "Per-Observation Silhouettes")
abline(h = 0, lty = 2)



ggplot(LDA_K,
       aes(k, avg_silhouette)) +
  geom_line() +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "Average Silhouette Width by Number of Clusters",
    x = "Number of Clusters",
    y = "Average Silhouette Width"
    )


certainty <- apply(Fuzzy_k$U, 1, max)

ggplot(
  data.frame(
    Obs = 1:length(certainty),
    Certainty = certainty,
    Cluster = factor(Fuzzy_Clust)
  ),
  aes(Obs, Certainty, colour = Cluster)
  ) +
  geom_hline(
    yintercept = 1/5,
    linetype = "dashed",
    colour = "red"
  ) +
  labs(
    caption = "Red dashed line indicates complete ambiguity (1/k)"
    ) +
  geom_point(size = 6) +
  theme_minimal() +
  labs(
    title = "Fuzzy Membership Certainty",
    y = "Maximum Membership"
    )



#not plotting right variables, fixed above.

#par(mar = c(6,5,4,6.5)) #stuff to help make the legend



#plot(Fuzzy_Clust, pch = 15, 
     #col = my_col[apply(Fuzzy_k$U, 1, which.max)],
     #xlab = "Obs",
     #ylab = "Silhouette Width",
     #main = "Fuzzy Silhouette Plot Using k-Medoids")
#legend("topright",
       #inset = c(-0.15,0),
       #legend = c("1", "2", "3", "4", "5","6"),
       #fill = c("#648FFF", "#785EF0", "#DC267F", "#FE6100","#FFB000","#F0E442"), xpd = TRUE)

#on.exit((par(old_par))) #stuff to help make the legend

```

The graph of Fuzzy Membership Certainty shows that the majority of our observations are near ambiguous in there assigned groupings.

## 5. Results

@Caponetti2017-fd determined that the quantity of sample was attributed to error generation. Both samples S38 and S08 had respective quantities 12.3mg and 16.0mg. The average amount of material sampled was reported at \~20mg.

The original study only had access to 38/40 ingots as they were on loan to another institute. This is an extension of that analysis, increasing the sample size at this point of the data analysis would inflate the false positive rate.

## 6. Discussion

@Caponetti2017-sc determined 3 distinct groupings of ingots, which provides evidence of a scarce and highly prized alloy being manufactured in one place at one time. This was determined with the added variable of the physical form of the ingots.

Our results

The initial analysis was performed by non-invasive XRF, a method of getting accurate readings within a certain range of elements (find that bit again). The ICP-OE and ICP-MS chemometric techniques require a small amount of material to be sacrificed in order to obtain a full range of the elemental composition of each ingot.

The find was dated to \~600BCE, the approximate age of the ship, and by the pottery also found in the wreck.

A study on a sample group this large has never been done before. Also, this dataset is not very large. - love this sentence 10/10.

XRF is a portable, non-invasive method of getting accurate readings within a certain range of elements (find that bit again). The ICP-OE and ICP-MS chemometric techniques require a small amount of material to be sacrificed in order to obtain a full range of the elemental composition of each ingot.

@Caponetti2017-fd performed chemoanalysis to determine the trace elements of these orichalcum ingots. Using this data, they determined correlation between the observed different features and the composition to group the ingots into their appropriate cluster. This is an extension from a fledgling study by the same author, where they performed a less invasive XRF (X-Ray Fluorescence) spectroscopy of the ingots, a non-distructive test that can be performed in the field.

Their motivation was to create a starting point for giving insight into archaeometric study of the orichalcum ingots regarding their origin and manufacture technologies.

Our study also has the same motivations. We aim to provide a high level data analysis to provide a detailed, and logical, path to a linkage analysis.

Motivation, problem statement, approach, results, conclusions.

What were the limitations of this study?

What further research could be done? Different statistical analysis?

When working with museum quality pieces, that need destructive analyses performed to

Data obtained for the (NAME TABLE) containing the main component, trace element concentrations and Cu/Zn weight ratio was obtained by ICP-OES and ICP-MS. Collecting samples from 38 ancient ingots to analysis the association between the ingots elemental composition.

We also took a preliminary look at the Copper (Cu), Zinc (Zn), Lead (Pb), and Tin (Sn) as they are the four alloying elements that are known to define orichalcum and historically have been the primary elemental components used to date and pinpoint origin of other orichalcum artifacts.

## 7. Refrences

::: {#refs}
:::
