---
title: "Re do"
author: "Gaia Smith"
date: "2026-08-31"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##  Packages to load/install:
```{r packages, include=FALSE}

library(tabulapdf)
library(dplyr)
library(tidyr)
library(pdftools)
library(ggcorrplot)
library(mvtnorm)
library(GGally)


```
##  Reading the pdf:
```{r reading pdf, echo=TRUE}

#Heres how to load it you will have to write a path to where you keep the pdf on your computer:
ingots <- pdf_text("C:/Users/smith/Desktop/STAT394/Group Project/A multivariate approach to the study of orichalcum ingots from the underwater Gela's archaeological site.pdf")

#check you have all the pages
length(ingots) #should out put 8

#prints page 3 which has the table on:
cat(ingots[[3]])

#this works if we want to view the pdf! not extract the data

```

##  Extracting the table from the pdf:
```{r extracting the table, echo=TRUE}

ingots2 <- extract_tables("C:/Users/smith/OneDrive/Desktop/A multivariate approach to the study of orichalcum ingots from the underwater Gela's archaeological site.pdf", pages=3)

head(ingots2)

print(ingots2)

#lets store it in a table:

table1 <- ingots2[[1]]

#when I fist this this you couldn't view Ni, Ag and Sb separately they all went under concentration
View(table1)

```

##  Cleaning the table:
```{r cleaning the table, echo=TRUE}
table1 <- table1 %>%
  separate(
    `Concentration in PPM`,
    into = c("Ni", "Ag", "Sb"),
    sep = " "
  )

#how to check:
table1[1:3, ] |> 
  print(width = Inf)

#this should now have all the columns separate :)
View(table1)

#making sure all the columns are named correctly:
names(table1) <- c(
  "Sample", "Cu", "Zn", "Pb", "Fe",
  "Ni", "Ag", "Sb", "As", "Co", "Cd", "Sn",
  "Te", "Bi", "Mn", "Li", "Al", "V", "Cr",
  "Rb", "Sr", "Ba", "Cu_Zn"
)

View(table1)

```


##  Creating a Data Frame of the Concentrations:
```{r making a new table, echo=TRUE}

Cu <- as.numeric(table1$Cu[-1])
Zn <- as.numeric(table1$Zn[-1])
Pb <- as.numeric(table1$Pb[-1])
Fe <- as.numeric(table1$Fe[-1])
Ni <- as.numeric(table1$Ni[-1])
Ag <- as.numeric(table1$Ag[-1])
Sb <- as.numeric(table1$Sb[-1])
As <- as.numeric(table1$As[-1])
Co <- as.numeric(table1$Co[-1])
Cd <- as.numeric(table1$Cd[-1])
Sn <- as.numeric(table1$Sn[-1])
Te <- as.numeric(table1$Te[-1])
Bi <- as.numeric(table1$Bi[-1])
Mn <- as.numeric(table1$Mn[-1])
Li <- as.numeric(table1$Li[-1])
Al <- as.numeric(table1$Al[-1])
V <- as.numeric(table1$V[-1])
Cr <- as.numeric(table1$Cr[-1])
Rb <- as.numeric(table1$Rb[-1])
Sr <- as.numeric(table1$Sr[-1])
Ba <- as.numeric(table1$Ba[-1])
Cu_Zn <- as.numeric(table1$Cu_Zn[-1])

Ingots_Frame <- data.frame (Sample = c("S1","S2","S3","S4","S5","S6","S7","S8","S9","S10","S11","S12","S13","S14","S15","S16","S17","S18","S19","S20","S21","S22","S23","S24","S25","S26","S27","S28","S29","S30","S31","S32","S33","S34","S35","S36","S37","S38"),Cu, Zn, Pb, Fe, Ni, Ag, Sb, As, Co, Cd, Sn, Te, Bi, Mn, Li, Al, V, Cr, Rb, Sr, Ba, Cu_Zn)

View(Ingots_Frame)

```

##  Summary of the raw data & Basic boxplot (no transformations)
```{r basic summary & boxplot, echo=TRUE}

#Summary of the Data Frame:
summary(Ingots_Frame)

max(Ingots_Frame[,-1])

#Basic boxplot of the raw data: (no transformations)
boxplot(Ingots_Frame[,-1], notch = TRUE, col="lightblue")

```
Notes:
Our raw data of concentrations produced a summary and simple box plots from above some things to note are Ni (nikel) has the largest spread followed by Sb (tin). Unfortunately due to the fact the data has a large range of concentrates (ranging from a minimum of 0.06 for Rb and a max of 683 for Ni) we will have to consider some transformations to allow for this massive difference in concentrations.
looking at the raw data we also noticed some interesting outliers particularly the 549 for sample 13 for Sb (tin), which the original paper noted as well.

We will be taking a deeper dive into Copper, Zinc, Lead and Tin as ... {something}

## Deeper summary and box plot of Cu, Zn, Pb and Sn:
```{r basic summary & boxplot, echo=TRUE}

#Summary of the Data Frame:
summary(Ingots_Frame[,c(2,3,4,12)])


#Basic boxplot of the raw data: (no transformations)
boxplot(Ingots_Frame[,c(2,3,4,12)], notch = TRUE, col="lightblue")

```
From the these boxplots we can more clearly see differences between spreads between our main elements, Lead and Zinc have a slight spread where as Copper and tine have a larger spread, Tin also has some outliers, particularly 65.8 from sample 17, as well as 42 from sample 22. Again Transformations will give us a better picture to show the relationships between the variable 

##  QQ-normal plots: (taking the log)
```{r qqnormal, echo=TRUE}


#no transformations
qqnorm(as.matrix(Ingots_Frame[-1,-1]), col="lightgray")
qqline(as.matrix(Ingots_Frame[-1,-1]), col="red")

#taking the log
qqnorm(log(as.matrix(Ingots_Frame[-1,-1])), col="darkgray")
qqline(log(as.matrix(Ingots_Frame[-1,-1])), col="red")

```

##: Getting MAD to work:
```{r transformations, echo=TRUE}
#Try one:
Mad_Ingots <- sapply(Ingots_Frame[,-1],mad)

print(Mad_Ingots)

#try two:
#Calculate medians and MADs for each column:
medians<- apply(Ingots_Frame[,-1], 2, median)
mads<- apply(Ingots_Frame[,-1], 2, mad)

medians
mads


#Compute the robust z-scores:
robust_z <- scale(Ingots_Frame[,-1], center = medians, scale = mads)

#Flag observations where any variable exceeds the threshold (e.g., 3)
outliers <- apply(abs(robust_z) > 3, 1, any)

# View the rows flagged as outliers
Ingots_Frame[outliers, ]

boxplot(robust_z)


qqnorm((robust_z), col="gray")
qqline((robust_z), col="red")


```



## Correlation and Covariance tables:
```{r basic correlation, echo=TRUE}

#Covariance table of the data:
var(Ingots_Frame[-1,-1])


#Correlation table of the data:
cor(Ingots_Frame[-1,-1])


#Correlation plot:
ggcorrplot(cor(Ingots_Frame[,-1]),
           method = "circle",
           hc.order = TRUE,
           type = "lower")


```


## Pairwise plot between Cu, Zn, Pb, and Sn:
```{r pairwise, echo=TRUE}

ggpairs(log(Ingots_Frame[,c(2,3,4,12)]))

```





















##  Playing with the Data (you can ignore this):
```{r playing, echo=TRUE}

#summary of the Data Frame:
summary(Ingots_Frame)


#Correlation plot
ggcorrplot(cor(Ingots_Frame[,-1]),
           method = "circle",
           hc.order = TRUE,
           type = "lower")

#Basic boxplot (taking the log):
boxplot(log(Ingots_Frame[,-1]), notch = TRUE, col="green")

boxplot(mad(Ingots_Frame[,-1]), notch = TRUE, col="purple")

boxplot(mad(Ingots_Frame[,-1], center = median(Ingots_Frame[,-1]), constant = 1.4826, na.rm = FALSE, low = FALSE, high = FALSE))

boxplot(sapply(Ingots_Frame[,-1], mad, na.rm = TRUE))


boxplot(mad(Ingots_Frame$Cu), mad(Ingots_Frame$Zn))

```

##  Covariance matrix
```{r covariance matrix, echo=TRUE}
Cov_Ingots<-cov(Ingots_Frame[,-1])
print(Cov_Ingots)

```


## Mahalanobis distance stuff, and QQ plot:
```{r Maha, echo=TRUE}

Maha_Ingots <- mahalanobis(Ingots_Frame[,-1], colMeans(Ingots_Frame[,-1]), var(Ingots_Frame[,-1]))

print(Maha_Ingots)

pchisq(Maha_Ingots, df=22)

plot(Maha_Ingots)

#Mahalanobis plot:

Maha_Ingots2 <- mahalanobis(Ingots_Frame[,-1], colMeans(Ingots_Frame[,-1]), cov(Ingots_Frame[,-1]))

upper.quantiles <- qchisq(c(.9, .95, .99), df=22)
density.at.quantiles <- dchisq(x=upper.quantiles, df=22)
cut.points <- data.frame(upper.quantiles, density.at.quantiles)

ggplot(data.frame(Maha_Ingots2), aes(x=Maha_Ingots2)) +
  geom_histogram(aes(y=after_stat(density)), bins=nclass.FD(Maha_Ingots2),
fill="white", col="black") +
  stat_function(fun=dchisq, args = list(df=13),
col="blue", linewidth=2, alpha=.7, xlim=c(0,65)) +
  geom_point(aes(x=Maha_Ingots, y=0), size=3, col="red") +
  geom_segment(data=cut.points,
aes(x=upper.quantiles, xend=upper.quantiles,
y=rep(0,3), yend=density.at.quantiles),
col="red", linewidth=1) + xlab("Mahalanobis distances and cut points") +
ylab("Histogram and density")


#QQ plot: with Malabouls distance
qqplot(qchisq(ppoints(1000), df=22), Maha_Ingots, main = expression("Q-Q Plot of Mahalanobis"* ~Maha_Ingots^2* "vs. quantiles of" * ~ chi[1]^2))
abline(0,1, col="red")
```


