---
title: "Data Analysis of Orichalcum Ingots"
author: "Gaia Smith"
date: "2026-08-31"
output: github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##  Packages to load/install:
```{r packages, include=FALSE}

library(tabulapdf)
library(dplyr)
library(tidyr)
library(pdftools)
library(ggcorrplot)
library(mvtnorm)
library(GGally)
library(here)
library(tinytex)
library(httr)
library(MVN)
library(ggplot2)
library(cluster)
library(factoextra)
library(hexbin)
library(matrixStats)
library(fclust)
```
##  Reading the pdf: - pdftools package code
```{r reading pdf, echo=TRUE}

#Load paper by caponnetti et. al from repository and store locally temporarily

url <- "https://raw.githubusercontent.com/luke8471/STAT394-GROUP-3/02d02760e329f00dd8cc8cc8543327ffeccd00a6/Text/Articles/caponetti%20et%20al.pdf"

tmp <- tempfile(fileext = ".pdf")

pdf <- httr::GET(url, write_disk(tmp, overwrite = TRUE))

ingots <- pdftools::pdf_text(tmp)


#check you have all the pages
length(ingots) #should out put 8

#prints page 3 which has the table on:
#cat(ingots[[3]])

#this works if we want to view the pdf! not extract the data

```

##  Extracting the table from the pdf: - tabulapdf package codes
```{r extracting the table, echo=TRUE}

ingots2 <- extract_tables(url, pages=3)

head(ingots2)

#lets store it in a table:

table1 <- ingots2[[1]]

#when I first did this this you couldn't view Ni, Ag and Sb separately they all went under concentration
```

##  Cleaning the table: - DATA202 codes
```{r cleaning the table, echo=TRUE}
table1 <- table1 %>%
  separate(
    'Concentration in PPM',
    into = c("Ni", "Ag", "Sb"),
    sep = " "
  )

#this should now have all the columns separate :)
#View(table1)

#making sure all the columns are named correctly:
names(table1) <- c(
  "Sample", "Cu", "Zn", "Pb", "Fe",
  "Ni", "Ag", "Sb", "As", "Co", "Cd", "Sn",
  "Te", "Bi", "Mn", "Li", "Al", "V", "Cr",
  "Rb", "Sr", "Ba", "Cu_Zn"
)

head(table1)
```


##  Creating a Data Frame of the Concentrations:
```{r making a new table, echo=TRUE}

Cu <- as.numeric(table1$Cu[-1])
Zn <- as.numeric(table1$Zn[-1])
Pb <- as.numeric(table1$Pb[-1])
Fe <- as.numeric(table1$Fe[-1])
Ni <- as.numeric(table1$Ni[-1])
Ag <- as.numeric(table1$Ag[-1])
Sb <- as.numeric(table1$Sb[-1])
As <- as.numeric(table1$As[-1])
Co <- as.numeric(table1$Co[-1])
Cd <- as.numeric(table1$Cd[-1])
Sn <- as.numeric(table1$Sn[-1])
Te <- as.numeric(table1$Te[-1])
Bi <- as.numeric(table1$Bi[-1])
Mn <- as.numeric(table1$Mn[-1])
Li <- as.numeric(table1$Li[-1])
Al <- as.numeric(table1$Al[-1])
V <- as.numeric(table1$V[-1])
Cr <- as.numeric(table1$Cr[-1])
Rb <- as.numeric(table1$Rb[-1])
Sr <- as.numeric(table1$Sr[-1])
Ba <- as.numeric(table1$Ba[-1])
Cu_Zn <- as.numeric(table1$Cu_Zn[-1])

Ingots_Frame <- data.frame (Sample = c("S1","S2","S3","S4","S5","S6","S7","S8","S9","S10","S11","S12","S13","S14","S15","S16","S17","S18","S19","S20","S21","S22","S23","S24","S25","S26","S27","S28","S29","S30","S31","S32","S33","S34","S35","S36","S37","S38"),Cu, Zn, Pb, Fe, Ni, Ag, Sb, As, Co, Cd, Sn, Te, Bi, Mn, Li, Al, V, Cr, Rb, Sr, Ba, Cu_Zn)

head(Ingots_Frame)

# Removing Cu/Zn from the table
Ingots_F_no <- Ingots_Frame[,1:22]

head(Ingots_F_no)

```

*Note:*
Had to do this as R was having a hard time working from the table 1
=======
##  Summary of the raw data (no transformations)
```{r basic summary, echo=TRUE}

#Summary of the Data Frame:
summary(Ingots_F_no)


#OLD Basic boxplot of the raw data: (no transformations)
#boxplot(Ingots_Frame[,-1], notch = TRUE, col="lightblue")

#Nicer box plot of the initial data: (no transformations) using ggplot:
#start by turing the data into long format:
Ingots_long <- Ingots_F_no[,-1] |>
  pivot_longer(
    cols = everything(),
    names_to = "Elements",
    values_to = "Concentrations")



ggplot(Ingots_long, aes(x= Elements, y= Concentrations)) +
  geom_boxplot(notch = FALSE, fill ="#40B0A6") +
  theme_minimal()+
  labs(title = "Intial boxplot of the raw data (no transformations)")




#boxplot(Ingots_F_no_Cu_zn[,-1], notch = TRUE, col="lightgreen")

#Covariance table of the data:
var(Ingots_F_no[-1,-1])


#Correlation table of the data:
cor(Ingots_F_no[-1,-1])


#Correlation plot:
ggcorrplot(cor(Ingots_F_no[,-1]),
           hc.order = TRUE,
           type = "lower",
           lab = TRUE,
           lab_size = 2,
           ggtheme = theme_minimal(),
           colors = c("#5D3A9B", "white","#E66100"),
           tl.col = "gray30",
           title = "Correlation plot of the raw data (no transformations)")

```

*Notes:*
Our raw data of concentrations produced a summary and simple box plots from above some things to note are Ni (nickel) has the largest spread followed by Sb (tin). Unfortunately due to the fact the data has a large range of concentrates (ranging from a minimum of 0.06 for Rb and a max of 683 for Ni) we will have to consider some transformations to allow for this massive difference in concentrations.
looking at the raw data we also noticed some interesting outliers particularly the 549 for sample 13 for Sb (tin), which the original paper noted as well.

We will be taking a deeper dive into Copper, Zinc, Lead and Tin as ... {something - Alice got this}




##  Deeper summary and box plot of Cu, Zn, Pb and Sn:
```{r basic summary & boxplot, echo=TRUE}

#Summary of the Data Frame:
summary(Ingots_F_no[,c(2,3,4,12)])


#Basic boxplot of the raw data: (no transformations)
boxplot(Ingots_F_no[,c(2,3,4,12)], notch = TRUE, col="lightblue")


```

*Notes:*
From the these boxplots we can more clearly see differences between spreads between our main elements, Lead and Zinc have a slight spread where as Copper and tine have a larger spread, Tin also has some outliers, particularly 65.8 from sample 17, as well as 42 from sample 22. Again Transformations will give us a better picture to show the relationships between the variable 

##  QQ-normal plots: (taking the log) - code from STAT292
```{r qqnormal, echo=TRUE}
#no transformations

qqnorm(as.matrix(Ingots_F_no[-1,-1]), col="lightblue")
qqline(as.matrix(Ingots_F_no[-1,-1]), col="red")

#log transform data

logIngots <- log(as.matrix(Ingots_F_no[-1,-1]))
summary(logIngots)

#NOTE: we return a lot of negative values after a log transformation. Either need
#to add a constant to keep values non-negative or normalise data another way.
#This is where I start considering removing the massive outliers from the data set.

#taking the log anyway, for giggles.

qqnorm(logIngots, col = "#4169E1");
qqline(logIngots, col = "red")

#steps in data and long tails, not happy. Median Absolute Deviation (MAD) may be
#the answer. Very robust to outliers. AMENDMENT best option fo sho

#HZ test on OG data, which also rejects null.

HZoriginal <- mvn(data = Ingots_F_no, mvn_test = "hz")
cat("Henze_Zirkler Test Results:\n")
print(HZoriginal$multivariate_normality)

#steps in data and long tails, not happy. Median Absolute Deviation (MAD) may be
#the answer. Very robust to outliers. AMENDMENT best option fo sho

```
*Notes:*
Looking at the QQ-plot with the light blue points (no transformation) we can see some clear deviations in the tails (lower and mostly upper data points), a low amount of points follow the theoretical line (in red), we can assume that our normality assumption wouldn't hold. hence transformations are needed. Their are also some slight steps in the points as well as a large jump at 1.7. 

Performing log transformations on the data produces the QQ-plot with blue points. Notice there are slight deviations in the tails (lower and upper data point), but the rest of the points follow the theoretical line (in red), we can assume that our data  normality assumption holds.


##  Getting MAD to work: - code from somewhere online (https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/mad and https://www.geeksforgeeks.org/r-language/calculate-the-median-absolute-deviation-in-r-programming-mad-function/)
```{r transformations-mad, echo=TRUE}

#Calculate medians and MAD for each column:
median <- apply(Ingots_F_no[,-1], 2, median)
mad <- apply(Ingots_F_no[,-1], 2, mad)


print(median/mad)
mad

#looking at this we can see some MASSIVE deviations, as expected. Now use the
#robust Z score to identify outliers that are candidates for removal.

#In univariate statistics, the Median Absolute Deviation is the most robust dispersion/scale measure in presence of outliers, and hence we strongly recommend the median plus or minus 2.5 times the MAD method for outlier detection. (Leys et al 2013)


#Compute the robust z-scores:
z <- scale(Ingots_F_no[,-1], center = median, scale = mad)

#Flag observations where any variable exceeds the threshold
outliers <- apply(abs(z) > 2.5, 1, any)

list(outliers)

# View the rows flagged as outliers
summary(Ingots_F_no[outliers,])

#The QQ plot shows very skewed data (expected)

qqnorm((z), col="gray");
qqline((z), col="red")

num_cols <- sapply(Ingots_F_no, is.numeric)
z_scores <- as.data.frame(scale(Ingots_F_no[, num_cols]))

summary(z_scores)

ol <- Ingots_F_no[rowSums(abs(z_scores) > 4) > 0, ]

print(ol)

Ingots_clean <- Ingots_F_no[rowSums(abs(z_scores) > 3) == 0, ]

summary(Ingots_clean)

#old boxplot:
#boxplot(Ingots_clean[,-1], notch = TRUE, col="lightblue")

#New boxplot:
Ingots_clean_long <- Ingots_clean[,-1] |>
  pivot_longer(
    cols = everything(),
    names_to = "Elements",
    values_to = "Concentrations")



ggplot(Ingots_clean_long, aes(x= Elements, y= Concentrations)) +
  geom_boxplot(notch = FALSE, fill ="#40B0A6") +
  theme_minimal()+
  labs(title = "Boxplot of the clean data")


#still getting negative values (even with elevated magnitude cutoff) in here so 
#log transforming may not be appropriate.

logIngotsClean <- log(as.matrix(Ingots_clean[-1,-1]))
summary(logIngotsClean)

View(logIngotsClean)


#making the dataframe numeric so it can be manipulated.

num_vars <- sapply(Ingots_clean, is.numeric)

#We COULD do this for every variable, and maybe should for primary components.

#PAIRWISE COMPARISON?

ggplot(Ingots_clean, aes(sample = Cu)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  theme_minimal()

#Henze-Zirkler test on cleaned data?
#Non-negative functional distance that measures between two distribution functions. If data
#is MVN, the test statistic HZ is approximately lognormally distributed. It calculates mean,
#variance, and smoothness parameter. Mean and variance are then log-normalised and p-value
#is estimated.
#@MISC{noauthor_undated-de,
#  howpublished = "\url{https://www.rdocumentation.org/packages/MVN/versions/5.8/topics/hzTest}",
#  note         = "Accessed: 2026-9-15"
#}

HZ <- mvn(data = Ingots_clean, mvn_test = "hz")

cat("Henze_Zirkler Test Results:\n")
print(HZ$multivariate_normality)

#Reject null and conclude that at pvalue=0.05, this dataset deviates from multivariate normality


#I think at this point we reject MVN for the WHOLE dataset.

#consider log transforming the trace elements..

#We COULD do this for every variable, and maybe should for primary components.

ggplot(Ingots_clean, aes(sample = Cu)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  theme_minimal()

ggplot(Ingots_clean, aes(sample = Zn)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  theme_minimal()

ggplot(Ingots_clean, aes(sample = Sn)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  theme_minimal()

ggplot(Ingots_clean, aes(sample = Sb)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  theme_minimal()

ggplot(Ingots_clean, aes(sample = Ni)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  theme_minimal()

ggplot(Ingots_clean, aes(sample = Ag)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  theme_minimal()

ggplot(Ingots_clean, aes(sample = As)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  theme_minimal()

```

##  Correlation and Covariance tables: - STAT394 inference pdf
```{r correlation/covariance clean data, echo=TRUE}

#performed for tidied data too


var(Ingots_clean[-1,-1])

cor(Ingots_clean[-1,-1])

ggcorrplot(cor(Ingots_clean[,-1]),
           hc.order = TRUE,
           type = "lower",
           lab = TRUE,
           lab_size = 1.5,
           ggtheme = theme_minimal(),
           colors = c("#5D3A9B", "white","#E66100"),
           tl.col = "gray30",
           title = "Correlation plot of the clean data exluding Cu/Zn")


#seeing a lot more correlation in the data with the massive outliers removed.
#worth keeping and commenting on?

```

*Notes:*
Correlation plot some things to highlight: 
___*Strong positive association between V and Cr (0.95547839) as well as between Li and Cr (0.830284850). 
=======

boxplot(Ingots_Frame[,-1], notch = TRUE, col="lightblue")


___*Strong negative association between Cu/Zn and Zn (0.93177327) which makes sense due to Cu/Zn being a ratio that includes Zn. Due to this we will mostly be removing Cu/Zn to keep independence??? (Alice knows more about this)


##  Pairwise plot between Cu, Zn, Pb, and Sn: - STAT394 inference pdf
```{r pairwise, echo=TRUE}
#old graph: using the raw data
#ggpairs(Ingots_Frame[,c(2,3,4,12)])

#new graph: using the clean data
ggpairs(Ingots_clean, columns = c(2,3,4,12),
        lower = list(continuous = wrap("points", alpha = 0.6, size = 1.5, color = "#006CD1")),
        upper = list(continuous = wrap("cor", size = 4, color = "#005AB5")),
        diag = list(continuous = wrap("densityDiag", alpha = 0.5, color = "#40B0A6")))


#Also done to reflect the traditional primary components that make up orichalcum,
#Zn, Cu, 

```

##  PCA stuff:
```{r PCA, echo=TRUE}
#Playing around with some ideas for PCA with common components in orichalcum.
#Tomorrow I will do some more research into the primary elements becauase it seems
#like it will be more useful. These make up common markers for origin etc. when
#archaeometallurgists are trying to identify casting method / intentional impurities / 
#where ores were mines from etc. 

#I think a good way to interpret and present this dataset is to segregate it in a
#way that makes sense to the people who will read and use this data for its 
#intended purpose, which is to compare it to known combinations of metals to try
#and find the ore source and how/where it was cast.

pca_data <- Ingots_clean[, c("Cu", "Zn", "Pb", "Fe", "Ni", "Ag", "Sb", "As")]

pca <- prcomp(pca_data, center = TRUE, scale. = TRUE)


biplot(pca, main = "PCA of only a few elements from the clean data")

#trying to see how it looks with all of them:

pca_all <- Ingots_clean[-1,-1]
#View(pca_all)
pca_1 <- prcomp(pca_all, center = TRUE, scale = TRUE)


biplot(pca_1, main = "PCA of all elements in the clean data")

#ggbiplot(pca_all, obs.scale = 1, var.scale = 1,groups = Sample$Group,elipse = TRUE,)


```

## LDA

```{r LDA, echo=TRUE}
#PAM (Partitioning Around Medoids) analysis 
#Trying with both OG and clean data
#Manhattan or Eucledian?

#Ingots_clean
#Ingots_clean_no (no CU/Zn)
#Ingots_Frame

ICN <- sapply(Ingots_clean[,-1], as.numeric)

scaled <- scale(ICN)

scaled

head(scaled, n = 3)

#need to find optimal silhouette width

sil_avg <- nrow(Ingots_clean[,-1])

for (k in 2:26) {
  PAM_fit <- pam(scaled, k)
  sil_avg[k - 1] <- PAM_fit$silinfo$avg.width
    }

data.frame(k = 2:26, avg_silhouette = sil_avg)



#The silhouette value/metric is a measurement of the quality of a clustering.
#between -1 and 1

#Reports optimal silhouette width as 5 with 0.31014624, indicating some weak 
#clustering... But it's the best we got.

ICN1 <- sapply(Ingots_clean[,-1], as.numeric)

scaled1 <- scale(ICN1)

scaled1

head(scaled1, n = 3)

sil_avg1 <- nrow(Ingots_clean[,-1])

for (k in 2:26) {
  PAM_fit <- pam(scaled, k)
  sil_avg[k - 1] <- PAM_fit$silinfo$avg.width
    }

data.frame(k = 2:26, avg_silhouette = sil_avg)


#using complete dataset or any of the other variations does not affect the
#suggested average silhoutte. Stick with 5. ***6 is also very close


pam <- pam(Ingots_clean[,-1], k = 5)

cluster <- pam$clustering

summary(pam)

fviz_silhouette(pam)+
  scale_fill_manual( values = c("#648FFF", "#785EF0", "#DC267F", "#FE6100","#FFB000"))+
  scale_colour_manual( values = c("#648FFF", "#785EF0", "#DC267F", "#FE6100","#FFB000"))
  

```

##  Cluster plot of the LDA stuff: - using code from the STAT394 cluster analysis notes (09, in NUKU)
```{r cluster plot, echo=TRUE}



result <- kmeans(ICN, 
                 algorithm = "MacQueen",
                 centers = 5, iter.max = 100, nstart = 50)

print(result)

#trying to do the clusterplot
factoextra::fviz_cluster(result, data = ICN,
                         palette = c("#648FFF", "#785EF0", "#DC267F", "#FE6100","#FFB000"),
                         ellipse.type = "euclid", 
                         star.plot = TRUE,
                         repel = TRUE, 
                         ggtheme = theme_minimal())


```


##  Hcluster: trying
```{r hclustering, echo=TRUE}

wards <- hclust(dist(scaled), method = "ward.D")
plot(wards)
rect.hclust(wards, 5, border = c("#648FFF", "#785EF0", "#DC267F", "#FE6100","#FFB000"))

#based on Alice's PAM we are using k = 5

```

## Trying stuff

```{r}
library(umap)

k <- umap(scaled)

UMAP_plot <- data.frame(
  UMAP1 = k$layout[,1],
  UMAP2 = k$layout[,2],
  Cluster = factor(PamMan$cluster)
  )

print(UMAP_plot)

ggplot(plot,
       aes(UMAP1, UMAP2,
           colour = Cluster)) +
  geom_point(size = 5) +
  stat_ellipse(level = 0.95)+
  scale_colour_manual( values = c("#648FFF", "#785EF0", "#DC267F", "#FE6100","#FFB000"))+
  labs(title = "Clustering of the clean data")

UMAP <- umap(scaled, metric = "manhattan")

head(UMAP)

PamMan <- pam(scaled, k = 5, metric = "manhattan")

plotPamMan <- data.frame(
  UMAP1 = umap$layout[,1],
  UMAP2 = umap$layout[,2],
  Cluster = factor(PamMan$cluster)
  )

ggplot(plotPamMan, 
       aes(UMAP1, UMAP2, color = cluster, size = medoid)) +
  geom_point(size = 5) +
  theme_classic() +
  labs(title = "UMAP of Data Using Manhattan Distance")
               
```

## more test

```{r}
#manhattan distance instead. More robust to outliers than Eucledian. We have big boi
#outliers and our clustering looks awful tbh lol.

dist <- dist(scaled, method = "manhattan")

HiMan <- hclust(dist, method = "average")

plot(HiMan)

rect.hclust(HiMan , 5, border = c("#648FFF", "#785EF0", "#DC267F", "#FE6100","#FFB000"))

```


## Fuzzy Analysis

```{r}

#scaled1

#k-medoids for Fuzzy

Fuzzy_k <- FKM.med(scaled, k = 6, m = 2)

#silhouettes

print(Fuzzy_k$U)

Fuzzy_Sil <- SIL.F(scaled1, Fuzzy_k$U)

Fuzzy_Clust <- apply(Fuzzy_k$U, 1, which.max)

table(Fuzzy_Clust)

plot(Fuzzy_Clust, pch = 15, 
     col = apply(Fuzzy_k$U, 1, which.max),
     xlab = "Obs",
     ylab = "Silhouette Width",
     main = "Fuzzy Silhouette Plot Using k-Medoids")

```




