---
title: "Re do"
author: "Gaia Smith"
date: "2026-08-31"
output:
  pdf_document: default
  html_document: default
output: output=github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##  Packages to load/install:
```{r packages, include=FALSE}

library(tabulapdf)
library(dplyr)
library(tidyr)
library(pdftools)
library(ggcorrplot)
library(mvtnorm)
library(GGally)


```
##  Reading the pdf: - pdftools package code
```{r reading pdf, echo=TRUE}

#Heres how to load it you will have to write a path to where you keep the pdf on your computer:
ingots <- pdf_text("C:/Users/smith/Desktop/STAT394/Group Project/A multivariate approach to the study of orichalcum ingots from the underwater Gela's archaeological site.pdf")

#check you have all the pages
length(ingots) #should out put 8

#prints page 3 which has the table on:
cat(ingots[[3]])

#this works if we want to view the pdf! not extract the data

```

##  Extracting the table from the pdf: - tabulapdf package codes
```{r extracting the table, echo=TRUE}

ingots2 <- extract_tables("C:/Users/smith/OneDrive/Desktop/A multivariate approach to the study of orichalcum ingots from the underwater Gela's archaeological site.pdf", pages=3)

head(ingots2)

print(ingots2)

#lets store it in a table:

table1 <- ingots2[[1]]

#when I fist this this you couldn't view Ni, Ag and Sb separately they all went under concentration
View(table1)

```

##  Cleaning the table: - DATA202 codes
```{r cleaning the table, echo=TRUE}
table1 <- table1 %>%
  separate(
    `Concentration in PPM`,
    into = c("Ni", "Ag", "Sb"),
    sep = " "
  )

#how to check:
table1[1:3, ] |> 
  print(width = Inf)

#this should now have all the columns separate :)
View(table1)

#making sure all the columns are named correctly:
names(table1) <- c(
  "Sample", "Cu", "Zn", "Pb", "Fe",
  "Ni", "Ag", "Sb", "As", "Co", "Cd", "Sn",
  "Te", "Bi", "Mn", "Li", "Al", "V", "Cr",
  "Rb", "Sr", "Ba", "Cu_Zn"
)

View(table1)

```


##  Creating a Data Frame of the Concentrations:
```{r making a new table, echo=TRUE}

Cu <- as.numeric(table1$Cu[-1])
Zn <- as.numeric(table1$Zn[-1])
Pb <- as.numeric(table1$Pb[-1])
Fe <- as.numeric(table1$Fe[-1])
Ni <- as.numeric(table1$Ni[-1])
Ag <- as.numeric(table1$Ag[-1])
Sb <- as.numeric(table1$Sb[-1])
As <- as.numeric(table1$As[-1])
Co <- as.numeric(table1$Co[-1])
Cd <- as.numeric(table1$Cd[-1])
Sn <- as.numeric(table1$Sn[-1])
Te <- as.numeric(table1$Te[-1])
Bi <- as.numeric(table1$Bi[-1])
Mn <- as.numeric(table1$Mn[-1])
Li <- as.numeric(table1$Li[-1])
Al <- as.numeric(table1$Al[-1])
V <- as.numeric(table1$V[-1])
Cr <- as.numeric(table1$Cr[-1])
Rb <- as.numeric(table1$Rb[-1])
Sr <- as.numeric(table1$Sr[-1])
Ba <- as.numeric(table1$Ba[-1])
Cu_Zn <- as.numeric(table1$Cu_Zn[-1])

Ingots_Frame <- data.frame (Sample = c("S1","S2","S3","S4","S5","S6","S7","S8","S9","S10","S11","S12","S13","S14","S15","S16","S17","S18","S19","S20","S21","S22","S23","S24","S25","S26","S27","S28","S29","S30","S31","S32","S33","S34","S35","S36","S37","S38"),Cu, Zn, Pb, Fe, Ni, Ag, Sb, As, Co, Cd, Sn, Te, Bi, Mn, Li, Al, V, Cr, Rb, Sr, Ba, Cu_Zn)

View(Ingots_Frame)

```

<<<<<<< HEAD
*Note:*
Had to do this as R was having a hard time working from the table 1
=======
##  Summary of the raw data (no transformations)
```{r basic summary, echo=TRUE}

#Summary of the Data Frame:
summary(Ingots_Frame)


#Basic boxplot of the raw data: (no transformations)
boxplot(Ingots_Frame[,-1], notch = TRUE, col="lightblue")


#Covariance table of the data:
var(Ingots_Frame[-1,-1])


#Correlation table of the data:
cor(Ingots_Frame[-1,-1])


#Correlation plot:
ggcorrplot(cor(Ingots_Frame[,-1]),
           method = "circle",
           hc.order = TRUE,
           type = "lower")


```

*Notes:*
Our raw data of concentrations produced a summary and simple box plots from above some things to note are Ni (nikel) has the largest spread followed by Sb (tin). Unfortunately due to the fact the data has a large range of concentrates (ranging from a minimum of 0.06 for Rb and a max of 683 for Ni) we will have to consider some transformations to allow for this massive difference in concentrations.
looking at the raw data we also noticed some interesting outliers particularly the 549 for sample 13 for Sb (tin), which the original paper noted as well.

We will be taking a deeper dive into Copper, Zinc, Lead and Tin as ... {something - Alice got this}




##  Deeper summary and box plot of Cu, Zn, Pb and Sn:
```{r basic summary & boxplot, echo=TRUE}

#Summary of the Data Frame:
summary(Ingots_Frame[,c(2,3,4,12)])


#Basic boxplot of the raw data: (no transformations)
boxplot(Ingots_Frame[,c(2,3,4,12)], notch = TRUE, col="lightblue")

```
*Notes:*
From the these boxplots we can more clearly see differences between spreads between our main elements, Lead and Zinc have a slight spread where as Copper and tine have a larger spread, Tin also has some outliers, particularly 65.8 from sample 17, as well as 42 from sample 22. Again Transformations will give us a better picture to show the relationships between the variable 

##  QQ-normal plots: (taking the log) - code from STAT292
```{r qqnormal, echo=TRUE}


#no transformations
qqnorm(as.matrix(Ingots_Frame[-1,-1]), col="lightblue")
qqline(as.matrix(Ingots_Frame[-1,-1]), col="red")

#taking the log
qqnorm(log(as.matrix(Ingots_Frame[-1,-1])), col="blue")
qqline(log(as.matrix(Ingots_Frame[-1,-1])), col="red")

```
*Notes:*
Looking at the QQ-plot with the light blue points (no transformation) we can see some clear deviations in the tails (lower and mostly upper data points), a low amount of points follow the theoretical line (in red), we can assume that our normality assumption wouldn't hold. hence transformations are needed. Their are also some slight steps in the points as well as a large jump at 1.7. 

Performing log transformations on the data produces the QQ-plot with blue points. Notice there are slight deviations in the tails (lower and upper data point), but the rest of the points follow the theoretical line (in red), we can assume that our data  normality assumption holds.




##  Getting MAD to work: - code from somewhere online (https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/mad and https://www.geeksforgeeks.org/r-language/calculate-the-median-absolute-deviation-in-r-programming-mad-function/)
```{r transformations, echo=TRUE}
#Try one:
Mad_Ingots <- sapply(Ingots_Frame[,-1],mad)

print(Mad_Ingots)

#try two:
#Calculate medians and MADs for each column:
medians<- apply(Ingots_Frame[,-1], 2, median)
mads<- apply(Ingots_Frame[,-1], 2, mad)

medians
mads


#Compute the robust z-scores:
robust_z <- scale(Ingots_Frame[,-1], center = medians, scale = mads)

#Flag observations where any variable exceeds the threshold (e.g., 3)
outliers <- apply(abs(robust_z) > 3, 1, any)

# View the rows flagged as outliers
Ingots_Frame[outliers, ]

boxplot(robust_z)


qqnorm((robust_z), col="gray")
qqline((robust_z), col="red")


```

##  Correlation and Covariance tables: - STAT394 inference pdf
```{r basic correlation, echo=TRUE}

#Covariance table of the data:
var(Ingots_Frame[-1,-1])


#Correlation table of the data:
cor(Ingots_Frame[-1,-1])

#Correlation plot:
ggcorrplot(cor(Ingots_Frame[,-1]),
           method = "circle",
           hc.order = TRUE,
           type = "lower")


<<<<<<< HEAD
```
*Notes:*
Correlation plot some things to highlight: 
___*Strong positive association between V and Cr (0.95547839) as well as between Li and Cr (0.830284850). 
=======

boxplot(Ingots_Frame[,-1], notch = TRUE, col="lightblue")
>>>>>>> 023f10788811bde61fd4c6fba27f95bf5ce6c2a8

___*Strong negative association between Cu/Zn and Zn (0.93177327) which makes sense due to Cu/Zn being a ratio that includes Zn. Due to this we will mostly be removing Cu/Zn to keep independence??? (Alice knows more about this)

##  looking closer at the at the strong positive correlation between V and Cr: - STAT394 inference pdf
```{r V and CR, echo=TRUE}
#3D static version of the density between V and Cr
mu1 <- colMeans(Ingots_Frame[,18:19])
sigma1 <- var(Ingots_Frame[,18:19])

n <-38
x <- seq(mu1[1]-3*sqrt(sigma1[1,1]), mu1[1]+3*sqrt(sigma1[1,1]), len=n)
y <- seq(mu1[2]-3*sqrt(sigma1[2,2]), mu1[2]+3*sqrt(sigma1[2,2]), len=n)
xy <- expand.grid(x=x, y=x)
z <- dmvnorm(xy, mu1, sigma1)
V.and.CR <- matrix(z, nrow = n)
persp(x, y, V.and.CR)

#contour curves for V and Cr:
n1 <- 100
x1 <- seq(mu1[1]-4*sqrt(sigma1[1,1]), mu1[1]+4*sqrt(sigma1[1,1]), len=n)
y1 <- seq(mu1[2]-4*sqrt(sigma1[2,2]), mu1[2]+4*sqrt(sigma1[2,2]), len=n)
xy1 <- expand.grid(x=x1, y=y1)
q.samp <- cbind(xy1, prob=dmvnorm(xy1, mean = mu1, sigma = sigma1))

p05 <- qmvnorm(0.5, mean=mu1, sigma = sigma1, tail = "both")$quantile
p09 <- qmvnorm(0.9, mean=mu1, sigma = sigma1, tail = "both")$quantile
p095 <- qmvnorm(0.95, mean=mu1, sigma = sigma1, tail = "both")$quantile
p099 <- qmvnorm(0.99, mean=mu1, sigma = sigma1, tail = "both")$quantile

dp <- c(dmvnorm(c(p05, p05), mean=mu1, sigma = sigma1),
        dmvnorm(c(p09, p09), mean=mu1, sigma = sigma1),
        dmvnorm(c(p095, p095), mean=mu1, sigma = sigma1),
        dmvnorm(c(p099, p099), mean=mu1, sigma = sigma1))
ggplot(q.samp, aes(x=x1, y=y1, z=prob)) + 
  geom_contour(breaks=dp) +
  coord_fixed(xlim = c(mu1[1]-4*sqrt(sigma1[1,1]), mu1[1]+4*sqrt(sigma1[1,1])), ylim = c(mu1[2]-4*sqrt(sigma1[2,2]), mu1[2]+4*sqrt(sigma1[2,2]))) 





```

##  looking closer at the at the strong negtive association between Cu/Zn and Zn: - STAT394 inference pdf
```{r Cu/Zn and Zn, echo=TRUE}
mu2 <- colMeans(Ingots_Frame[,c(2,23)])
sigma2 <- var(Ingots_Frame[,c(2,23)])

n <-38
v <- seq(mu2[1]-3*sqrt(sigma2[1,1]), mu2[1]+3*sqrt(sigma2[1,1]), len=n)
w <- seq(mu2[2]-3*sqrt(sigma2[2,2]), mu2[2]+3*sqrt(sigma2[2,2]), len=n)
vw <- expand.grid(x=v, y=w)
q <- dmvnorm(vw, mu2, sigma2)
Cu_Zn.and.Zn <- matrix(q, nrow = n)
persp(x, y, Cu_Zn.and.Zn)

#contour curves for V and Cr:
n2 <- 100
x2 <- seq(mu2[1]-4*sqrt(sigma2[1,1]), mu2[1]+4*sqrt(sigma2[1,1]), len=n)
y2 <- seq(mu2[2]-4*sqrt(sigma2[2,2]), mu2[2]+4*sqrt(sigma2[2,2]), len=n)
xy2 <- expand.grid(x=x2, y=y2)
q.samp_2 <- cbind(xy2, prob=dmvnorm(xy2, mean = mu2, sigma = sigma2))

p05_2 <- qmvnorm(0.5, mean=mu2, sigma = sigma2, tail = "both")$quantile
p09_2 <- qmvnorm(0.9, mean=mu2, sigma = sigma2, tail = "both")$quantile
p095_2 <- qmvnorm(0.95, mean=mu2, sigma = sigma2, tail = "both")$quantile
p099_2 <- qmvnorm(0.99, mean=mu2, sigma = sigma2, tail = "both")$quantile

dp_2 <- c(dmvnorm(c(p05_2, p05_2), mean=mu2, sigma = sigma2),
        dmvnorm(c(p09_2, p09_2), mean=mu2, sigma = sigma2),
        dmvnorm(c(p095_2, p095_2), mean=mu2, sigma = sigma2),
        dmvnorm(c(p099_2, p099_2), mean=mu2, sigma = sigma2))
ggplot(q.samp_2, aes(x=x2, y=y2, z=prob)) + 
  geom_contour(breaks=dp_2) +
  coord_fixed(
    xlim = c(mu2[1]-4*sqrt(sigma2[1,1]), mu2[1]+4*sqrt(sigma2[1,1])),
    ylim = c(mu2[2]-4*sqrt(sigma2[2,2]), mu2[2]+4*sqrt(sigma2[2,2]))) 



```



##  Pairwise plot between Cu, Zn, Pb, and Sn: - STAT394 inference pdf
```{r pairwise, echo=TRUE}

ggpairs(log(Ingots_Frame[,c(2,3,4,12)]))

```

##  Mahalanobis distance stuff, and QQ plot: - STAT394 inference pdf
```{r Maha, echo=TRUE}

Maha_Ingots <- mahalanobis(Ingots_Frame[,-1], colMeans(Ingots_Frame[,-1]), var(Ingots_Frame[,-1]))

pchisq(Maha_Ingots, df=22) #using 22 df as we have 22 concentrations minus (Cn/Zn)


#Mahalanobis plot:

Maha_Ingots2 <- mahalanobis(Ingots_Frame[,-1], colMeans(Ingots_Frame[,-1]), cov(Ingots_Frame[,-1]))

upper.quantiles <- qchisq(c(.9, .95, .99), df=22)
density.at.quantiles <- dchisq(x=upper.quantiles, df=22)
cut.points <- data.frame(upper.quantiles, density.at.quantiles)

ggplot(data.frame(Maha_Ingots2), aes(x=Maha_Ingots2)) +
  geom_histogram(aes(y=after_stat(density)), bins=nclass.FD(Maha_Ingots2),
fill="white", col="black") +
  stat_function(fun=dchisq, args = list(df=13),
col="blue", linewidth=2, alpha=.7, xlim=c(0,65)) +
  geom_point(aes(x=Maha_Ingots, y=0), size=3, col="red") +
  geom_segment(data=cut.points,
aes(x=upper.quantiles, xend=upper.quantiles,
y=rep(0,3), yend=density.at.quantiles),
col="red", linewidth=1) + xlab("Mahalanobis distances and cut points") +
ylab("Histogram and density")


<<<<<<< HEAD
#QQ plot: with Malabouls distance
qqplot(qchisq(ppoints(1000), df=22), Maha_Ingots, main = expression("Q-Q Plot of Mahalanobis"* ~Maha_Ingots^2* "vs. quantiles of" * ~ chi[1]^2))
abline(0,1, col="red")
```

=======
#QQ plot:
qqplot(qchisq(ppoints(1000), df=22), log(Maha_Ingots), main = expression("Q-Q Plot of Mahalanobis"* ~Maha_Ingots^2* "vs. quantiles of" * ~ chi[1]^2))
abline(0,1, col="red")
```


## QQ normal plot
```{r normal, echo=TRUE}
qqnorm(as.data.frame(Ingots_Frame[-1,-1]))



```


##  Transformations:
```{r transformations, echo=TRUE}
>>>>>>> 023f10788811bde61fd4c6fba27f95bf5ce6c2a8



##  Playing with the Data (you can ignore this): - Gaia
```{r playing, echo=TRUE}

#summary of the Data Frame:
summary(Ingots_Frame)


#Correlation plot
ggcorrplot(cor(Ingots_Frame[,-1]),
           method = "circle",
           hc.order = TRUE,
           type = "lower")

#Basic boxplot (taking the log):
boxplot(log(Ingots_Frame[,-1]), notch = TRUE, col="green")

boxplot(mad(Ingots_Frame[,-1]), notch = TRUE, col="purple")

boxplot(mad(Ingots_Frame[,-1], center = median(Ingots_Frame[,-1]), constant = 1.4826, na.rm = FALSE, low = FALSE, high = FALSE))

boxplot(sapply(Ingots_Frame[,-1], mad, na.rm = TRUE))


boxplot(mad(Ingots_Frame$Cu), mad(Ingots_Frame$Zn))

#log of our data set:
ggplot(Ingots_Frame) +
   geom_boxplot()

ggplot(log(Ingots_Frame[,-1]), aes(x = c(), y = Sample))
       geom_boxplot()




```

