---
title: "Re do"
author: "Gaia Smith"
date: "2026-08-31"
output: github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##  Packages to load/install:
```{r packages, include=FALSE}

library(tabulapdf)
library(dplyr)
library(tidyr)
library(pdftools)
library(ggcorrplot)
library(mvtnorm)
library(GGally)
library(here)
library(tinytex)
library(httr)
library(MVN)
library(ggplot2)
```
##  Reading the pdf: - pdftools package code
```{r reading pdf, echo=TRUE}

#Load paper by caponnetti et. al from repository and store locally temporarily

url <- "https://raw.githubusercontent.com/luke8471/STAT394-GROUP-3/02d02760e329f00dd8cc8cc8543327ffeccd00a6/Text/Articles/caponetti%20et%20al.pdf"

tmp <- tempfile(fileext = ".pdf")

pdf <- httr::GET(url, write_disk(tmp, overwrite = TRUE))

ingots <- pdftools::pdf_text(tmp)


#check you have all the pages
length(ingots) #should out put 8

#prints page 3 which has the table on:
cat(ingots[[3]])

#this works if we want to view the pdf! not extract the data

```

##  Extracting the table from the pdf: - tabulapdf package codes
```{r extracting the table, echo=TRUE}

ingots2 <- extract_tables(url, pages=3)

head(ingots2)

print(ingots2)

#lets store it in a table:

table1 <- ingots2[[1]]

#when I first did this this you couldn't view Ni, Ag and Sb separately they all went under concentration
View(table1)

```

##  Cleaning the table: - DATA202 codes
```{r cleaning the table, echo=TRUE}
table1 <- table1 %>%
  separate(
    'Concentration in PPM',
    into = c("Ni", "Ag", "Sb"),
    sep = " "
  )

#how to check:
table1[1:3, ] |> 
  print(width = Inf)

#this should now have all the columns separate :)
View(table1)

#making sure all the columns are named correctly:
names(table1) <- c(
  "Sample", "Cu", "Zn", "Pb", "Fe",
  "Ni", "Ag", "Sb", "As", "Co", "Cd", "Sn",
  "Te", "Bi", "Mn", "Li", "Al", "V", "Cr",
  "Rb", "Sr", "Ba", "Cu_Zn"
)

View(table1)

```


##  Creating a Data Frame of the Concentrations:
```{r making a new table, echo=TRUE}

Cu <- as.numeric(table1$Cu[-1])
Zn <- as.numeric(table1$Zn[-1])
Pb <- as.numeric(table1$Pb[-1])
Fe <- as.numeric(table1$Fe[-1])
Ni <- as.numeric(table1$Ni[-1])
Ag <- as.numeric(table1$Ag[-1])
Sb <- as.numeric(table1$Sb[-1])
As <- as.numeric(table1$As[-1])
Co <- as.numeric(table1$Co[-1])
Cd <- as.numeric(table1$Cd[-1])
Sn <- as.numeric(table1$Sn[-1])
Te <- as.numeric(table1$Te[-1])
Bi <- as.numeric(table1$Bi[-1])
Mn <- as.numeric(table1$Mn[-1])
Li <- as.numeric(table1$Li[-1])
Al <- as.numeric(table1$Al[-1])
V <- as.numeric(table1$V[-1])
Cr <- as.numeric(table1$Cr[-1])
Rb <- as.numeric(table1$Rb[-1])
Sr <- as.numeric(table1$Sr[-1])
Ba <- as.numeric(table1$Ba[-1])
Cu_Zn <- as.numeric(table1$Cu_Zn[-1])

Ingots_Frame <- data.frame (Sample = c("S1","S2","S3","S4","S5","S6","S7","S8","S9","S10","S11","S12","S13","S14","S15","S16","S17","S18","S19","S20","S21","S22","S23","S24","S25","S26","S27","S28","S29","S30","S31","S32","S33","S34","S35","S36","S37","S38"),Cu, Zn, Pb, Fe, Ni, Ag, Sb, As, Co, Cd, Sn, Te, Bi, Mn, Li, Al, V, Cr, Rb, Sr, Ba, Cu_Zn)

View(Ingots_Frame)

# Removing Cu/Zn from the table to see if it changes stuff mostly for independence and just to see what happens
Ingots_F_no_Cu_zn <- Ingots_Frame[,1:22]

View(Ingots_F_no_Cu_zn)

```

*Note:*
Had to do this as R was having a hard time working from the table 1
=======
##  Summary of the raw data (no transformations)
```{r basic summary, echo=TRUE}

#Summary of the Data Frame:
summary(Ingots_Frame)


#Basic boxplot of the raw data: (no transformations)
boxplot(Ingots_Frame[,-1], notch = TRUE, col="lightblue")

#boxplot(Ingots_F_no_Cu_zn[,-1], notch = TRUE, col="lightgreen")

#Covariance table of the data:
var(Ingots_Frame[-1,-1])


#Correlation table of the data:
cor(Ingots_Frame[-1,-1])


#Correlation plot:
ggcorrplot(cor(Ingots_Frame[,-1]),
           method = "circle",
           hc.order = TRUE,
           type = "lower")

ggcorrplot(cor(Ingots_F_no_Cu_zn[,-1]),
           method = "circle",
           hc.order = TRUE,
           type = "lower")


```

*Notes:*
Our raw data of concentrations produced a summary and simple box plots from above some things to note are Ni (nickel) has the largest spread followed by Sb (tin). Unfortunately due to the fact the data has a large range of concentrates (ranging from a minimum of 0.06 for Rb and a max of 683 for Ni) we will have to consider some transformations to allow for this massive difference in concentrations.
looking at the raw data we also noticed some interesting outliers particularly the 549 for sample 13 for Sb (tin), which the original paper noted as well.

We will be taking a deeper dive into Copper, Zinc, Lead and Tin as ... {something - Alice got this}




##  Deeper summary and box plot of Cu, Zn, Pb and Sn:

```{r basic summary & boxplot, echo=TRUE}

#Summary of the Data Frame:
summary(Ingots_Frame[,c(2,3,4,12)])


#Basic boxplot of the raw data: (no transformations)
boxplot(Ingots_Frame[,c(2,3,4,12)], notch = TRUE, col="lightblue")

```

*Notes:*
From the these boxplots we can more clearly see differences between spreads between our main elements, Lead and Zinc have a slight spread where as Copper and tine have a larger spread, Tin also has some outliers, particularly 65.8 from sample 17, as well as 42 from sample 22. Again Transformations will give us a better picture to show the relationships between the variable 

##  QQ-normal plots: (taking the log) - code from STAT292
```{r qqnormal, echo=TRUE}
#no transformations

qqnorm(as.matrix(Ingots_Frame[-1,-1]), col="lightblue")
qqline(as.matrix(Ingots_Frame[-1,-1]), col="red")

#log transform data

logIngots <- log(as.matrix(Ingots_Frame[-1,-1]))
summary(logIngots)

#NOTE: we return a lot of negative values after a log transformation. Either need
#to add a constant to keep values non-negative or normalise data another way.
#This is where I start considering removing the massive outliers from the data set.

#taking the log anyway, for giggles.

qqnorm(logIngots, col = "blue");
qqline(logIngots, col = "red")

#steps in data and long tails, not happy. Median Absolute Deviation (MAD) may be
#the answer. Very robust to outliers. AMENDMENT best option fo sho

#HZ test on OG data, which also rejects null.

HZoriginal <- mvn(data = Ingots_Frame, mvn_test = "hz")
cat("Henze_Zirkler Test Results:\n")
print(HZoriginal$multivariate_normality)




#all of the above but without Cu/Zn:
qqnorm(as.matrix(Ingots_F_no_Cu_zn[-1,-1]), col="lightgreen")
qqline(as.matrix(Ingots_F_no_Cu_zn[-1,-1]), col="red")

#log transform data

logIngots_no <- log(as.matrix(Ingots_F_no_Cu_zn[-1,-1]))
summary(logIngots_no)


qqnorm(logIngots_no, col = "green");
qqline(logIngots_no, col = "red")

#steps in data and long tails, not happy. Median Absolute Deviation (MAD) may be
#the answer. Very robust to outliers. AMENDMENT best option fo sho

#HZ test on OG data, which also rejects null.

HZoriginal_no <- mvn(data = Ingots_F_no_Cu_zn, mvn_test = "hz")
cat("Henze_Zirkler Test Results:\n")
print(HZoriginal_no$multivariate_normality)






```
*Notes:*
Looking at the QQ-plot with the light blue points (no transformation) we can see some clear deviations in the tails (lower and mostly upper data points), a low amount of points follow the theoretical line (in red), we can assume that our normality assumption wouldn't hold. hence transformations are needed. Their are also some slight steps in the points as well as a large jump at 1.7. 

Performing log transformations on the data produces the QQ-plot with blue points. Notice there are slight deviations in the tails (lower and upper data point), but the rest of the points follow the theoretical line (in red), we can assume that our data  normality assumption holds.




##  Getting MAD to work: - code from somewhere online (https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/mad and https://www.geeksforgeeks.org/r-language/calculate-the-median-absolute-deviation-in-r-programming-mad-function/)
```{r transformations-mad, echo=TRUE}
#Try one:
#Mad_Ingots <- sapply(Ingots_Frame[,-1],mad)

#print(Mad_Ingots)

#try two:
#Calculate medians and MADs for each column:
medians <- apply(Ingots_Frame[,-1], 2, median)
mads <- apply(Ingots_Frame[,-1], 2, mad)

print(medians/mads)
mads

#looking at this we can see some MASSIVE deviations, as expected. Now use the
#robust Z score to identify outliers that are candidates for removal.

#In univariate statistics, the Median Absolute Deviation is the most robust dispersion/scale measure in presence of outliers, and hence we strongly recommend the median plus or minus 2.5 times the MAD method for outlier detection. (Leys et al 2013)


#Compute the robust z-scores:
z <- scale(Ingots_Frame[,-1], center = medians, scale = mads)

#Flag observations where any variable exceeds the threshold
outliers <- apply(abs(z) > 2.5, 1, any)

list(outliers)

# View the rows flagged as outliers
summary(Ingots_Frame[outliers,])

#The QQ plot shows very skewed data (expected)

qqnorm((z), col="gray");
qqline((z), col="red")


#sorry this is an absolute mess, was trying a few methods. Need to tidy up

num_cols <- sapply(Ingots_Frame, is.numeric)
z_scores <- as.data.frame(scale(Ingots_Frame[, num_cols]))

summary(z_scores)

ol <- Ingots_Frame[rowSums(abs(z_scores) > 4) > 0, ]

print(ol)

Ingots_clean <- Ingots_Frame[rowSums(abs(z_scores) > 3) == 0, ]

summary(Ingots_clean)

boxplot(Ingots_clean[,-1], notch = TRUE, col="lightblue")

#still getting negative values (even with elevated magnitude cutoff) in here so 
#log transforming may not be appropriate.

logIngotsClean <- log(as.matrix(Ingots_clean[-1,-1]))
summary(logIngotsClean)


#making the dataframe numeric so it can be manipulated.

num_vars <- sapply(Ingots_clean, is.numeric)

#We COULD do this for every variable, and maybe should for primary components.

ggplot(Ingots_clean, aes(sample = Cu)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  theme_minimal()

#Henze-Zirkler test on cleaned data?
#Non-negative functional distance that measures between two distribution functions. If data
#is MVN, the test statistic HZ is approximately lognormally distributed. It calculates mean,
#variance, and smoothness parameter. Mean and variance are then log-normalised and p-value
#is estimated.
#@MISC{noauthor_undated-de,
#  howpublished = "\url{https://www.rdocumentation.org/packages/MVN/versions/5.8/topics/hzTest}",
#  note         = "Accessed: 2026-9-15"
#}

HZ <- mvn(data = Ingots_clean, mvn_test = "hz")

cat("Henze_Zirkler Test Results:\n")
print(HZ$multivariate_normality)

#Reject null and conclude that at pvalue=0.05, this dataset deviates from multivariate normality


#I think at this point we reject MVN for the WHOLE dataset.

#consider log transforming the trace elements..


##all of the above but without Cu/Zn:
medians_no <- apply(Ingots_F_no_Cu_zn[,-1], 2, median)
mads_no <- apply(Ingots_F_no_Cu_zn[,-1], 2, mad)

print(medians_no/mads_no)
mads_no


#Compute the robust z-scores:
z_no <- scale(Ingots_F_no_Cu_zn[,-1], center = medians_no, scale = mads_no)

#Flag observations where any variable exceeds the threshold
outliers_no <- apply(abs(z) > 2.5, 1, any)

list(outliers_no)

# View the rows flagged as outliers
summary(Ingots_F_no_Cu_zn[outliers,])

#The QQ plot shows very skewed data (expected)

qqnorm((z), col="gray");
qqline((z), col="red")


#sorry this is an absolute mess, was trying a few methods. Need to tidy up

num_cols_no <- sapply(Ingots_F_no_Cu_zn, is.numeric)
z_scores_no <- as.data.frame(scale(Ingots_F_no_Cu_zn[, num_cols_no]))

summary(z_scores)

ol_no <- Ingots_F_no_Cu_zn[rowSums(abs(z_scores) > 3) > 0, ]

print(ol_no)

Ingots_clean_no <- Ingots_F_no_Cu_zn[rowSums(abs(z_scores) > 3) == 0, ]

summary(Ingots_clean_no)

boxplot(Ingots_clean_no[,-1], notch = TRUE, col="lightgreen")

#still getting negative values (even with elevated magnitude cutoff) in here so 
#log transforming may not be appropriate.

logIngotsClean_no <- log(as.matrix(Ingots_clean_no[-1,-1]))
summary(logIngotsClean_no)


#making the dataframe numeric so it can be manipulated.

num_vars_no <- sapply(Ingots_clean_no, is.numeric)

#We COULD do this for every variable, and maybe should for primary components.

ggplot(Ingots_clean_no, aes(sample = Cu)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  theme_minimal()


HZ_no <- mvn(data = Ingots_clean_no, mvn_test = "hz")

cat("Henze_Zirkler Test Results:\n")
print(HZ_no$multivariate_normality)



```

## Mahalanobis

```{r}
#next should be Mahalanobis, it functions better with multivariate than MAD and 
#can be used in conjunction.


#Amendment: skip entirely and go to PCA.




```




##  Correlation and Covariance tables: - STAT394 inference pdf
```{r basic correlation, echo=TRUE}

#Covariance table of the data:
var(Ingots_Frame[-1,-1])


#Correlation table of the data:
cor(Ingots_Frame[-1,-1])

#Correlation plot:
ggcorrplot(cor(Ingots_Frame[,-1]),
           method = "circle",
           hc.order = TRUE,
           type = "lower")

#performed for tidied data too


var(Ingots_clean[-1,-1])

cor(Ingots_clean[-1,-1])

ggcorrplot(cor(Ingots_clean[,-1]),
           method = "circle",
           hc.order = TRUE,
           type = "lower")


#seeing a lot more correlation in the data with the massive outliers removed.
#worth keeping and commenting on?

```

*Notes:*
Correlation plot some things to highlight: 
___*Strong positive association between V and Cr (0.95547839) as well as between Li and Cr (0.830284850). 
=======

boxplot(Ingots_Frame[,-1], notch = TRUE, col="lightblue")


___*Strong negative association between Cu/Zn and Zn (0.93177327) which makes sense due to Cu/Zn being a ratio that includes Zn. Due to this we will mostly be removing Cu/Zn to keep independence??? (Alice knows more about this)

##  looking closer at the at the strong positive correlation between V and Cr: - STAT394 inference pdf
(this is supported by the correlation plots)
```{r V and CR, echo=TRUE}
#3D static version of the density between V and Cr
mu1 <- colMeans(Ingots_Frame[,18:19])
sigma1 <- var(Ingots_Frame[,18:19])

n <-38
x <- seq(mu1[1]-3*sqrt(sigma1[1,1]), mu1[1]+3*sqrt(sigma1[1,1]), len=n)
y <- seq(mu1[2]-3*sqrt(sigma1[2,2]), mu1[2]+3*sqrt(sigma1[2,2]), len=n)
xy <- expand.grid(x=x, y=x)
z <- dmvnorm(xy, mu1, sigma1)
V.and.CR <- matrix(z, nrow = n)
persp(x, y, V.and.CR)

#contour curves for V and Cr:
n1 <- 100
x1 <- seq(mu1[1]-4*sqrt(sigma1[1,1]), mu1[1]+4*sqrt(sigma1[1,1]), len=n)
y1 <- seq(mu1[2]-4*sqrt(sigma1[2,2]), mu1[2]+4*sqrt(sigma1[2,2]), len=n)
xy1 <- expand.grid(x=x1, y=y1)
q.samp <- cbind(xy1, prob=dmvnorm(xy1, mean = mu1, sigma = sigma1))

p05 <- qmvnorm(0.5, mean=mu1, sigma = sigma1, tail = "both")$quantile
p09 <- qmvnorm(0.9, mean=mu1, sigma = sigma1, tail = "both")$quantile
p095 <- qmvnorm(0.95, mean=mu1, sigma = sigma1, tail = "both")$quantile
p099 <- qmvnorm(0.99, mean=mu1, sigma = sigma1, tail = "both")$quantile

dp <- c(dmvnorm(c(p05, p05), mean=mu1, sigma = sigma1),
        dmvnorm(c(p09, p09), mean=mu1, sigma = sigma1),
        dmvnorm(c(p095, p095), mean=mu1, sigma = sigma1),
        dmvnorm(c(p099, p099), mean=mu1, sigma = sigma1))
ggplot(q.samp, aes(x=x1, y=y1, z=prob)) + 
  geom_contour(breaks=dp) +
  coord_fixed(xlim = c(mu1[1]-4*sqrt(sigma1[1,1]), mu1[1]+4*sqrt(sigma1[1,1])), ylim = c(mu1[2]-4*sqrt(sigma1[2,2]), mu1[2]+4*sqrt(sigma1[2,2]))) 





```

##  looking closer at the at the strong negative association between Cu/Zn and Zn: - STAT394 inference pdf
```{r Cu/Zn and Zn, echo=TRUE}
mu2 <- colMeans(Ingots_Frame[,c(2,23)])
sigma2 <- var(Ingots_Frame[,c(2,23)])

n <-38
v <- seq(mu2[1]-3*sqrt(sigma2[1,1]), mu2[1]+3*sqrt(sigma2[1,1]), len=n)
w <- seq(mu2[2]-3*sqrt(sigma2[2,2]), mu2[2]+3*sqrt(sigma2[2,2]), len=n)
vw <- expand.grid(x=v, y=w)
q <- dmvnorm(vw, mu2, sigma2)
Cu_Zn.and.Zn <- matrix(q, nrow = n)
persp(x, y, Cu_Zn.and.Zn)

#contour curves for V and Cr:
n2 <- 100
x2 <- seq(mu2[1]-4*sqrt(sigma2[1,1]), mu2[1]+4*sqrt(sigma2[1,1]), len=n)
y2 <- seq(mu2[2]-4*sqrt(sigma2[2,2]), mu2[2]+4*sqrt(sigma2[2,2]), len=n)
xy2 <- expand.grid(x=x2, y=y2)
q.samp_2 <- cbind(xy2, prob=dmvnorm(xy2, mean = mu2, sigma = sigma2))

p05_2 <- qmvnorm(0.5, mean=mu2, sigma = sigma2, tail = "both")$quantile
p09_2 <- qmvnorm(0.9, mean=mu2, sigma = sigma2, tail = "both")$quantile
p095_2 <- qmvnorm(0.95, mean=mu2, sigma = sigma2, tail = "both")$quantile
p099_2 <- qmvnorm(0.99, mean=mu2, sigma = sigma2, tail = "both")$quantile

dp_2 <- c(dmvnorm(c(p05_2, p05_2), mean=mu2, sigma = sigma2),
        dmvnorm(c(p09_2, p09_2), mean=mu2, sigma = sigma2),
        dmvnorm(c(p095_2, p095_2), mean=mu2, sigma = sigma2),
        dmvnorm(c(p099_2, p099_2), mean=mu2, sigma = sigma2))
ggplot(q.samp_2, aes(x=x2, y=y2, z=prob)) + 
  geom_contour(breaks=dp_2) +
  coord_fixed(
    xlim = c(mu2[1]-4*sqrt(sigma2[1,1]), mu2[1]+4*sqrt(sigma2[1,1])),
    ylim = c(mu2[2]-4*sqrt(sigma2[2,2]), mu2[2]+4*sqrt(sigma2[2,2]))) 



```



##  Pairwise plot between Cu, Zn, Pb, and Sn: - STAT394 inference pdf
```{r pairwise, echo=TRUE}

ggpairs(log(Ingots_Frame[,c(2,3,4,12)]))

#Also done to reflect the traditional primary components that make up orichalcum,
#Zn, Cu, 

```

##  Mahalanobis distance stuff, and QQ plot: - STAT394 inference pdf
```{r Maha, echo=TRUE}

Maha_Ingots <- mahalanobis(Ingots_Frame[,-1], colMeans(Ingots_Frame[,-1]), var(Ingots_Frame[,-1]))

pchisq(Maha_Ingots, df=22) #using 22 df as we have 22 concentrations minus (Cn/Zn)


#Mahalanobis plot:

Maha_Ingots2 <- mahalanobis(Ingots_Frame[,-1], colMeans(Ingots_Frame[,-1]), cov(Ingots_Frame[,-1]))

upper.quantiles <- qchisq(c(.9, .95, .99), df=22)
density.at.quantiles <- dchisq(x=upper.quantiles, df=22)
cut.points <- data.frame(upper.quantiles, density.at.quantiles)

ggplot(data.frame(Maha_Ingots2), aes(x=Maha_Ingots2)) +
  geom_histogram(aes(y=after_stat(density)), bins=nclass.FD(Maha_Ingots2),
fill="white", col="black") +
  stat_function(fun=dchisq, args = list(df=13),
col="blue", linewidth=2, alpha=.7, xlim=c(0,65)) +
  geom_point(aes(x=Maha_Ingots, y=0), size=3, col="red") +
  geom_segment(data=cut.points,
aes(x=upper.quantiles, xend=upper.quantiles,
y=rep(0,3), yend=density.at.quantiles),
col="red", linewidth=1) + xlab("Mahalanobis distances and cut points") +
ylab("Histogram and density")


#QQ plot: with Malabouls distance
qqplot(qchisq(ppoints(1000), df=22), Maha_Ingots, main = expression("Q-Q Plot of Mahalanobis"* ~Maha_Ingots^2* "vs. quantiles of" * ~ chi[1]^2))
abline(0,1, col="red")
```


#QQ plot:
qqplot(qchisq(ppoints(1000), df=22), log(Maha_Ingots), main = expression("Q-Q Plot of Mahalanobis"* ~Maha_Ingots^2* "vs. quantiles of" * ~ chi[1]^2))
abline(0,1, col="red")
```


## QQ normal plot
```{r normal, echo=TRUE}
qqnorm(as.data.frame(Ingots_Frame[-1,-1]))

```


##  Playing with the Data (you can ignore this): - Gaia
```{r playing, echo=TRUE}

#summary of the Data Frame:
summary(Ingots_Frame)


#Correlation plot
ggcorrplot(cor(Ingots_Frame[,-1]),
           method = "circle",
           hc.order = TRUE,
           type = "lower")

#Basic boxplot (taking the log):
boxplot(log(Ingots_Frame[,-1]), notch = TRUE, col="green")

boxplot(mad(Ingots_Frame[,-1]), notch = TRUE, col="purple")

boxplot(mad(Ingots_Frame[,-1], center = median(Ingots_Frame[,-1]), constant = 1.4826, na.rm = FALSE, low = FALSE, high = FALSE))

boxplot(sapply(Ingots_Frame[,-1], mad, na.rm = TRUE))


boxplot(mad(Ingots_Frame$Cu), mad(Ingots_Frame$Zn))

#log of our data set:
ggplot(Ingots_Frame) +
   geom_boxplot()

ggplot(log(Ingots_Frame[,-1]), aes(x = c(), y = Sample))
       geom_boxplot()




```

##  PCA stuff:
```{r PCA, echo=TRUE}
#Playing around with some ideas for PCA with common components in orichalcum.
#Tomorrow I will do some more research into the primary elements becauase it seems
#like it will be more useful. These make up common markers for origin etc. when
#archaeometallurgists are trying to identify casting method / intentional impurities / 
#where ores were mines from etc. 

#I think a good way to interpret and present this dataset is to segregate it in a
#way that makes sense to the people who will read and use this data for its 
#intended purpose, which is to compare it to known combinations of metals to try
#and find the ore source and how/where it was cast.

pca_data <- Ingots_clean[, c("Cu", "Zn", "Pb", "Fe", "Ni", "Ag", "Sb", "As")]

pca <- prcomp(pca_data, center = TRUE, scale. = TRUE)


biplot(pca)

#trying to see how it looks with all of them:

pca_all <- Ingots_clean[-1,-1]
#View(pca_all)
pca_1 <- prcomp(pca_all, center = TRUE, scale = TRUE)


biplot(pca_1)


```

