---
title: "Data Analysis of Orichalcum Ingots"
author: "Gaia Smith"
date: "2026-08-31"
output: github_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages to load/install:

```{r packages and set up, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(tabulapdf)
library(dplyr)
library(tidyr)
library(pdftools)
library(ggcorrplot)
library(mvtnorm)
library(GGally)
library(here)
library(tinytex)
library(httr)
library(MVN)
library(ggplot2)
library(cluster)
library(factoextra)
library(hexbin)
library(matrixStats)
library(fclust)
library(knitr)
library(tidyverse)
library(skimr)
library(bibtex)
library(ggdendro)
library(dendextend)
library(gridExtra)
library(gt)
library(factoextra)
library(umap)
library(moments)
library(mclust)
library(reshape2)
library(ggrepel)

my_col <- c("#648FFF", "#785EF0", "#DC267F", "#FE6100","#FFB000")

##  Reading the pdf: - pdftools package code:

#Load paper by caponnetti et. al from repository and store locally temporarily

url <- "https://raw.githubusercontent.com/luke8471/STAT394-GROUP-3/02d02760e329f00dd8cc8cc8543327ffeccd00a6/Text/Articles/caponetti%20et%20al.pdf"

tmp <- tempfile(fileext = ".pdf")

pdf <- httr::GET(url, write_disk(tmp, overwrite = TRUE))

ingots <- pdftools::pdf_text(tmp)

```

## Extracting the table from the pdf: - tabulapdf package codes

```{r extracting the table, echo=TRUE}

ingots2 <- extract_tables(url, pages=3)

head(ingots2)

#lets store it in a table:

table1 <- ingots2[[1]]

#when I first did this this you couldn't view Ni, Ag and Sb separately they all went under concentration
```

## Cleaning the table: - DATA202 codes

```{r cleaning the table, echo=TRUE}
table1 <- table1 %>%
  separate(
    'Concentration in PPM',
    into = c("Ni", "Ag", "Sb"),
    sep = " "
  )

#this should now have all the columns separate :)
#View(table1)

#making sure all the columns are named correctly:
names(table1) <- c(
  "Sample", "Cu", "Zn", "Pb", "Fe",
  "Ni", "Ag", "Sb", "As", "Co", "Cd", "Sn",
  "Te", "Bi", "Mn", "Li", "Al", "V", "Cr",
  "Rb", "Sr", "Ba", "Cu_Zn"
)

head(table1)
```

## Creating a Data Frame of the Concentrations:

```{r making a new table, echo=TRUE}
#| fig-width: 10
#| fig-height: 6

##  Extracting the table from the pdf: - tabulapdf package codes:
ingots2 <- suppressMessages(extract_tables(url, pages=3))

#lets store it in a table:

table1 <- suppressMessages(ingots2[[1]])

#when I first did this this you couldn't view Ni, Ag and Sb separately they all went under concentration
##  Cleaning the table: - DATA202 codes
table1 <- suppressMessages(table1 %>%
                             separate(
                               'Concentration in PPM',
                               into = c("Ni", "Ag", "Sb"),
                               sep = " "
                             ))

#this should now have all the columns separate :)

#making sure all the columns are named correctly:
suppressMessages(names(table1) <- c(
  "Sample", "Cu", "Zn", "Pb", "Fe",
  "Ni", "Ag", "Sb", "As", "Co", "Cd", "Sn",
  "Te", "Bi", "Mn", "Li", "Al", "V", "Cr",
  "Rb", "Sr", "Ba", "Cu_Zn"
))

#as R was having a hard time using the table above we created a Data Frame of the Concentrations:

Cu <- as.numeric(table1$Cu[-1])
Zn <- as.numeric(table1$Zn[-1])
Pb <- as.numeric(table1$Pb[-1])
Fe <- as.numeric(table1$Fe[-1])
Ni <- as.numeric(table1$Ni[-1])
Ag <- as.numeric(table1$Ag[-1])
Sb <- as.numeric(table1$Sb[-1])
As <- as.numeric(table1$As[-1])
Co <- as.numeric(table1$Co[-1])
Cd <- as.numeric(table1$Cd[-1])
Sn <- as.numeric(table1$Sn[-1])
Te <- as.numeric(table1$Te[-1])
Bi <- as.numeric(table1$Bi[-1])
Mn <- as.numeric(table1$Mn[-1])
Li <- as.numeric(table1$Li[-1])
Al <- as.numeric(table1$Al[-1])
V <- as.numeric(table1$V[-1])
Cr <- as.numeric(table1$Cr[-1])
Rb <- as.numeric(table1$Rb[-1])
Sr <- as.numeric(table1$Sr[-1])
Ba <- as.numeric(table1$Ba[-1])
Cu_Zn <- as.numeric(table1$Cu_Zn[-1])

Ingots_Frame <- data.frame (Sample = c("S1","S2","S3","S4","S5","S6","S7","S8","S9","S10","S11","S12","S13","S14","S15","S16","S17","S18","S19","S20","S21","S22","S23","S24","S25","S26","S27","S28","S29","S30","S31","S32","S33","S34","S35","S36","S37","S38"),Cu, Zn, Pb, Fe, Ni, Ag, Sb, As, Co, Cd, Sn, Te, Bi, Mn, Li, Al, V, Cr, Rb, Sr, Ba, Cu_Zn)

# Removing Cu/Zn from the table
Ingots_F_no <- Ingots_Frame[,1:22]

knitr::kable(Ingots_F_no,
             caption = "Figure 1: Table 1 The main components and trace elements from Caponetti et al study",
             format = "pipe")

#How to save table an image

#tab <- gt(Ingots_Frame)

#gtsave(
  #tab,
  #"LOCAL DIRECTORY"
  #)

```

```{r basic summary, echo=FALSE}
#| fig-width: 10
#| fig-height: 6

#Summary of the Data Frame:
invisible(summary(Ingots_F_no))

#code to make the summary into a table
skim(Ingots_F_no[,-1]) |>
  select(-n_missing, -complete_rate)


#Making summary stats of table 1


Element_order <- colnames(as.data.frame(Ingots_F_no[, -1]))

Ingots_sum <- Ingots_F_no[, -1] |>
  pivot_longer(
    cols = everything(),
    names_to = "Element",
    values_to = "Concentration"
  ) |>
  mutate(
    Element = factor(Element, levels = Element_order)
  ) |>
  group_by(Element) |>
  summarise(
    Mean = mean(Concentration, na.rm = TRUE),
    SD = sd(Concentration, na.rm = TRUE),
    Min = min(Concentration, na.rm = TRUE),
    Q1 = quantile(Concentration, 0.25, na.rm = TRUE),
    Median = median(Concentration, na.rm = TRUE),
    Q3 = quantile(Concentration, 0.75, na.rm = TRUE),
    Max = max(Concentration, na.rm = TRUE),
    IQR = IQR(Concentration, na.rm = TRUE),
    Kurtosis = kurtosis(Concentration, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(
  Ingots_sum, 
  digits = 2,
  caption = "Summary Statistics of Table 1", 
  format = "pipe")
  
#box plot of the initial data: (no transformations) using ggplot:

Ingots_long <- Ingots_F_no[, -1] |>
  pivot_longer(
    cols = everything(),
    names_to = "Element",
    values_to = "Concentration"
    )

#notchbox
  
ggplot( 
  Ingots_long,
   aes(
     x = Element,
     y = Concentration
     )
  ) +
  geom_boxplot(
    notch = TRUE,
    fill = "#648FFF",
    colour = "grey30",
    linewidth = 0.7,
    outlier.shape = 21,
    outlier.fill = "#FE6100",
    outlier.colour = "grey30",
    outlier.stroke = 0.8,
    outlier.size = 3
  ) +
  geom_jitter(
    colour = "#785EF0",
    alpha = 0.1,
    width = 0.12,
    size = 2
  ) +
  labs(
    title = "Element Concentration Spread Between Samples",
    subtitle = "38 Orichalcum Ingots | Notched Boxplots | Highlighted Outliers",
    x = "Element",
    y = "Concentration (PPM)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 22
    ),
    plot.subtitle = element_text(
      colour = "grey30",
      size = 14
    ),
    plot.title.position = "plot",
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

#ggsave(
  #file.path("C:/Users/Alice/Documents/STAT394-GROUP-3/Figures", "notchbox.png"),
  #p,
  #width = 10,
  #height = 6,
  #dpi = 300
#)

```

## QQ-normal plots: - code from STAT292

```{r qqnormal, echo=TRUE}
#| fig-width: 10
#| fig-height: 6



#no transformations

qqnorm(as.matrix(Ingots_F_no[-1,-1]), col="#648FFF")
qqline(as.matrix(Ingots_F_no[-1,-1]), col="#DC267F")

#no transformations: elements in percentages:

qqnorm(as.matrix(Ingots_F_no[-1,2:5]), col="#648FFF", main = "Normal QQ-Plot for Primary Elements")
qqline(as.matrix(Ingots_F_no[-1,2:5]), col="#DC267F")
qqline(as.matrix(Ingots_F_no[-1,2:5]), col="red")

#png(
  #"C:/Users/Alice/Documents/STAT394-GROUP-3/Figures/qq_primary.png",
  #width = 3000,
  #height = 1800,
  #res = 600
#)

qqnorm(
  as.matrix(Ingots_F_no[-1, 2:5]),
  col = "#648FFF",
  main = "Normal QQ-Plot for Primary Elements"
)

qqline(
  as.matrix(Ingots_F_no[-1, 2:5]),
  col = "#DC267F"
)

dev.off()

#no transformations: elements in PPM:

qqnorm(as.matrix(Ingots_F_no[-1,6:22]), col="#648FFF", main = "Normal QQ-Plot for Trace Elements")
qqline(as.matrix(Ingots_F_no[-1,6:22]), col="#DC267F")

#png(  
  #"C:/Users/Alice/Documents/STAT394-GROUP-3/Figures/qq_trace.png",
  #width = 3000,
  #height = 1800,
  #res = 600
  #)

qqnorm(
  as.matrix(Ingots_F_no[-1, 2:5]),
  col = "#648FFF",
  main = "Normal QQ-Plot for Trace Elements"
)

qqline(
  as.matrix(Ingots_F_no[-1, 2:5]),
  col = "#DC267F"
)

dev.off()

#log transform data

logIngots <- log(as.matrix(Ingots_F_no[-1,-1]))
#print(logIngots)

knitr::kable(logIngots,
             caption = "Log Transformed Data",
             digits = 2,
             format = "pipe")

```

#NOTE: we return several zero-values after a log transformation. Either need #to add a constant to keep values non-negative or normalise data another way. #This is where I start considering removing the massive outliers from the data set.

#Use log(x + 1) to eliminate zero values. It works

```{r}
logIngots1 <- log(as.matrix(Ingots_F_no[-1,-1] + 1))

#log(x + 1)

qqnorm(logIngots1, col = "#648FFF");
qqline(logIngots1, col = "#DC267F")

#steps in data and long tails, not happy. Median Absolute Deviation (MAD) may be
#the answer. Very robust to outliers. AMENDMENT best option fo sho


#HZ test on the log(x + 1) transformed data


HZoriginal <- MVN::mvn(data = logIngots1, mvn_test = "hz")
cat("Henze_Zirkler Test Results:\n")
print(HZoriginal$multivariate_normality)


#steps in data and long tails, not happy despite HZ reporting normal, p-value 
#still close. Median Absolute Deviation (MAD) may be the answer. Very robust to 
#outliers.


element_order <- c(
  "Cu", "Zn", "Pb", "Fe",
  "Ag", "Al", "As", "Ba", "Bi",
  "Cd", "Co", "Cr", "Li", "Mn",
  "Ni", "Rb", "Sb", "Sn", "Sr",
  "Te", "V"
)

qq_long <- Ingots_F_no |>
  pivot_longer(
    cols = 2:22,
    names_to = "Element",
    values_to = "Value"
  )


qq_long$Element <- factor(
  qq_long$Element,
  levels = element_order
)

ggplot(qq_long, aes(sample = Value)) +
  stat_qq(colour = "#648FFF") +
  stat_qq_line(colour = "#DC267F") +
  facet_wrap(~ Element, 
             scales = "free",
             ncol = 3
             ) +
  theme_minimal()

#ggsave(
  #"C:/Users/Alice/Documents/STAT394-GROUP-3/Figures/QQ_all.png",
  #width = 12,
  #height = 20,
  #dpi = 600
#)
```

*Note:* Had to do this as R was having a hard time working from the table 1

\## Summary of the raw data (no transformations) \## MAD

```{r transformations-mad, echo=TRUE}
#| fig-width: 10
#| fig-height: 6


#Calculate medians and MAD for each column:
median <- apply(logIngots1[,-1], 2, median)
mad <- apply(logIngots1[,-1], 2, mad)


print(median/mad)

#looking at this we can see some MASSIVE deviations, as expected. Now use the
#robust Z score to identify outliers that are candidates for removal.

#In univariate statistics, the Median Absolute Deviation is the most robust dispersion/scale measure in presence of outliers, and hence we strongly recommend the median plus or minus 2.5 times the MAD method for outlier detection. (Leys et al 2013)


#Compute the robust z-scores:
z <- scale(logIngots1[,-1], center = median, scale = mad)

#Flag observations where any variable exceeds the threshold
outliers <- apply(abs(z) > 2.5, 1, any)

# View the rows flagged as outliers
summary(logIngots1[outliers,])

#The QQ plot shows very skewed data (expected)

qqnorm((z), col="gray");
qqline((z), col="red")

z_scores <- as.data.frame(scale(logIngots1[, ]))

summary(z_scores)

ol <- logIngots1[rowSums(abs(z_scores) > 2.5) > 0, ]

Ingots_clean <- logIngots1[rowSums(abs(z_scores) > 2.5) == 0, ]

#boxplot standardised with outliers removed:

Ingots_clean_long <- as.data.frame(Ingots_clean[,-1]) |>
  pivot_longer(
    cols = everything(),
    names_to = "Elements",
    values_to = "Concentrations")

#normbox <- 

ggplot(
  Ingots_clean_long,
  aes(
    x= Elements,
    y= Concentrations
    )
  ) +
  geom_boxplot(
    notch = TRUE,
    fill = "#648FFF",
    colour = "grey30",
    linewidth = 0.7,
    outlier.shape = 21,
    outlier.fill = "#FE6100",
    outlier.colour = "grey30",
    outlier.stroke = 0.8,
    outlier.size = 3
  ) +
  geom_jitter(
    colour = "#785EF0",
    alpha = 0.1,
    width = 0.12,
    size = 2
  ) +
  labs(
    title = "Element Concentration Spread Between Samples",
    subtitle = "Outliers Removed | Standardised | Highlighted Outliers",
    x = "Element",
    y = "Concentration"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 22
    ),
    plot.subtitle = element_text(
      colour = "grey30",
      size = 14
    ),
    plot.title.position = "plot",
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

#ggsave(
  #file.path("C:/Users/Alice/Documents/STAT394-GROUP-3/Figures", "notchboxnorm.png"),
  #normbox,
  #width = 10,
  #height = 6,
  #dpi = 600
 #)


#no zero values, LDA is go

logIngotsClean <- log(as.matrix(Ingots_clean[-1,-1]))
summary(logIngotsClean)

#making the dataframe numeric so it can be manipulated.

num_vars <- sapply(Ingots_clean, is.numeric)

#QQ-plots for Ingots_clean

Ingots_clean_df <- as.data.frame(Ingots_clean)

qq_long2 <- Ingots_clean_df |>
  pivot_longer(
    cols = 1:21,
    names_to = "Element",
    values_to = "Value"
  )


qq_long2$Element <- factor(
  qq_long2$Element,
  levels = element_order
)

ggplot(qq_long2, aes(sample = Value)) +
  stat_qq(colour = "#648FFF") +
  stat_qq_line(colour = "#DC267F") +
  facet_wrap(~ Element, 
             scales = "free", 
             ncol = 3 
    ) +
  theme_minimal()

#ggsave(
  #"C:/Users/Alice/Documents/STAT394-GROUP-3/Figures/QQ_allnorm.png",
  #width = 12,
  #height = 20,
  #dpi = 600
#)

#Henze-Zirkler test on cleaned data?
#Non-negative functional distance that measures between two distribution functions. If data
#is MVN, the test statistic HZ is approximately lognormally distributed. It calculates mean,
#variance, and smoothness parameter. Mean and variance are then log-normalised and p-value
#is estimated.
#@MISC{noauthor_undated-de,
#  howpublished = "\url{https://www.rdocumentation.org/packages/MVN/versions/5.8/topics/hzTest}",
#  note         = "Accessed: 2026-9-15"
#}

HZ <- MVN::mvn(data = Ingots_clean, mvn_test = "hz")

cat("Henze_Zirkler Test Results:\n")
print(HZ$multivariate_normality)

```

```{r basic summary, echo=TRUE}
#| fig-width: 10
#| fig-height: 6

#Covariance table of the data:
invisible(var(Ingots_F_no[-1,-1]))


#Correlation table of the data:
invisible(cor(Ingots_F_no[-1,-1]))


#Correlation plot:
cor1 <- ggcorrplot(
  cor(Ingots_F_no[,-1]),
  hc.order = FALSE,
  type = "lower",
  lab = TRUE,
  lab_size = 1.5,
  ggtheme = theme_minimal(),
  colors = c("#FE6100", "white","#648FFF"),
  tl.col = "gray30"
  ) +
  labs(
    title = "Correlation Matrix",
    subtitle = "Pearson Correlations | Raw Data"
    ) +
  theme(
    plot.subtitle = element_text(
    colour = "grey30"
    ))

ggsave(
  "C:/Users/Alice/Documents/STAT394-GROUP-3/Figures/correlation_matrix.png",
  plot = cor1,
  width = 12,
  height = 10,
  dpi = 600
)


#performed for clean data:
invisible(var(Ingots_clean[-1,-1]))

invisible(cor(Ingots_clean[-1,-1]))

cor2 <- ggcorrplot(
  cor(Ingots_clean),
  hc.order = FALSE,
  type = "lower",
  lab = TRUE,
  lab_size = 1.5,
  ggtheme = theme_minimal(),
  colors = c("#FE6100", "white","#648FFF"),
  tl.col = "gray30",
  ) +
  labs(
    title = "Correlation Matrix",
    subtitle = "Pearson Correlations | log-Transformed Data"
    ) +
  theme(
    plot.subtitle = element_text(
    colour = "grey30"
    ))

ggsave(
  "C:/Users/Alice/Documents/STAT394-GROUP-3/Figures/correlation_matrix_log.png",
  plot = cor2,
  width = 12,
  height = 10,
  dpi = 600
)

```

*Notes:* Our raw data of concentrations produced a summary and simple box plots from above some things to note are Ni (nickel) has the largest spread followed by Sb (tin). Unfortunately due to the fact the data has a large range of concentrates (ranging from a minimum of 0.06 for Rb and a max of 683 for Ni) we will have to consider some transformations to allow for this massive difference in concentrations. looking at the raw data we also noticed some interesting outliers particularly the 549 for sample 13 for Sb (tin), which the original paper noted as well.

We will be taking a deeper dive into Copper, Zinc, Lead and Tin as ... {something - Alice got this}

## Deeper summary and box plot of Cu, Zn, Pb and Sn:

```{r basic summary & boxplot, echo=TRUE}

#Summary of the Data Frame:
summary(Ingots_F_no[,c(2,3,4,12)])

skim(Ingots_F_no[,c(2,3,4,12)]) |>
  dplyr::select(-n_missing, -complete_rate)



#Basic boxplot of the raw data: (no transformations)
boxplot(Ingots_F_no[,c(2,3,4,12)], notch = TRUE, col="lightblue")


```

*Notes:* From the these boxplots we can more clearly see differences between spreads between some of our most prevalent elements, Lead and Zinc have a slight spread where as Copper and tine have a larger spread, Tin also has some outliers, particularly 65.8 from sample 17, as well as 42 from sample 22. Again Transformations will give us a better picture to show the relationships between the variable

*Notes:* Looking at the QQ-plot with the light blue points (no transformation) we can see some clear deviations in the tails (lower and mostly upper data points), a low amount of points follow the theoretical line (in red), we can assume that our normality assumption wouldn't hold. hence transformations are needed. Their are also some slight steps in the points as well as a large jump at 1.7.

Performing log transformations on the data produces the QQ-plot with blue points. Notice there are slight deviations in the tails (lower and upper data point), but the rest of the points follow the theoretical line (in red), we can assume that our data normality assumption holds.

## Correlation and Covariance tables: - STAT394 inference pdf

```{r correlation/covariance clean data, echo=TRUE}

#performed for tidied data too


var(Ingots_clean[-1,-1])

cor(Ingots_clean[-1,-1])

ggcorrplot(
  cor(Ingots_clean[,-1]), 
  hc.order = FALSE,
  type = "lower",
  lab = TRUE,
  lab_size = 1.5,
  ggtheme = theme_minimal(),
  colors = c("#FE6100", "white","#648FFF"),
  tl.col = "gray30"
  ) +
  labs(
    title = "Correlation Matrix",
    subtitle = "Pearson Correlations | Tidied Data"
    ) +
  theme(
    plot.subtitle = element_text(
    colour = "grey30"
    ))



#seeing a lot more correlation in the data with the massive outliers removed.
#worth keeping and commenting on?

```

*Notes:* Correlation plot some things to highlight: \_\_\_\*Strong positive association between V and Cr (0.95547839) as well as between Li and Cr (0.830284850). =======

boxplot(Ingots_Frame[,-1], notch = TRUE, col="lightblue")

\_\_\_\*Strong negative association between Cu/Zn and Zn (0.93177327) which makes sense due to Cu/Zn being a ratio that includes Zn. Due to this we will mostly be removing Cu/Zn to keep independence??? (Alice knows more about this)

## Pairwise plot between Cu, Zn, Pb, and Sn: - STAT394 inference pdf

```{r pairwise, echo=TRUE}
#old graph: using the raw data
#ggpairs(Ingots_Frame[,c(2,3,4,12)])

#new graph: using the clean data
ggpairs(Ingots_clean, columns = c(1,2,3,4),
        lower = list(continuous = wrap("points", alpha = 0.6, size = 1.5, color = "#006CD1")),
        upper = list(continuous = wrap("cor", size = 4, color = "#005AB5")),
        diag = list(continuous = wrap("densityDiag", alpha = 0.5, color = "#40B0A6")))


#Also done to reflect the traditional primary components that make up orichalcum,
#Zn, Cu, 

```

## LDA

```{r LDA, echo=TRUE}
#| fig-width: 10
#| fig-height: 6

#PAM (Partitioning Around Medoids) analysis 
#Trying with both OG and clean data
#Manhattan or Eucledian?

ICN <- as.data.frame(Ingots_clean[,-1])

ICN[] <- lapply(ICN, as.numeric)

scaled <- scale(ICN)

#need to find optimal silhouette width

sil_avg <- numeric(25)

for (k in 2:25) {
  PAM_fit <- pam(scaled, k)
  sil_avg[k - 1] <- PAM_fit$silinfo$avg.width
    }

LDA_K <- data.frame(k = 2:26, avg_silhouette = sil_avg)


pam_sum_tab <- data.frame()


for (k in 2:25) {
  
  PAM_fit <- cluster::pam(scaled, k)
  
  sizes <- table(PAM_fit$clustering)
  
  pam_sum_tab <- rbind(
    pam_sum_tab,
    data.frame(
      k = k,
      Avg_Silhouette = round(PAM_fit$silinfo$avg.width, 3),
      Smallest_Cluster = min(sizes),
      Largest_Cluster = max(sizes),
      Cluster_Sizes = paste(sizes, collapse = ", ")
    )
  )
}

knitr::kable(
pam_sum_tab,
caption = "PAM clustering summary across candidate values of k.",
align = "c"
)

pam <- pam(Ingots_clean, k =3)

cluster <- pam$clustering

for (k in 2:25) {
  PAM_fit <- pam(scaled, k)
  sil_avg[k - 1] <- PAM_fit$silinfo$avg.width
    }

LDA_K1 <- data.frame(k = 2:26, avg_silhouette = sil_avg)


fviz_silhouette(pam)+
  scale_fill_manual( values = c("#648FFF", "#DC267F", "#FFB000"))+
  scale_colour_manual( values = c("gray60", "gray60", "gray60"))
 
#Rebuilding to match others graphics

pam_fit <- pam(scaled, k = 3)
d <- dist(scaled, method = "manhattan")
pam_sil <- silhouette(pam_fit$clustering, d)
sil_df <- as.data.frame(pam_sil)

sil_df$Sample <- rownames(scaled)

sil_df <- sil_df |>
  dplyr::arrange(cluster, desc(sil_width)) |>
  dplyr::mutate(
    Observation = row_number()
  )

sil_df$cluster <- factor(sil_df$cluster)

sil_avg <- mean(sil_df$sil_width)

ggplot(sil_df,
  aes(
    x = Observation,
    y = sil_width,
    fill = cluster
  )
) +
  geom_col(width = 0.8) +
  geom_text(
    aes(label = Sample),
    angle = 0,
    size = 4,
    vjust = -0.3 
  ) +
  geom_hline(
    yintercept = sil_avg,
    colour = "red",
    linetype = "dashed"
  ) +
  scale_fill_manual(
    values = c(
      "#648FFF",
      "#DC267F",
      "#FFB000"
    )
  ) +
  labs(
    title = "PAM Cluster Silhouette Plot",
    subtitle = paste(
      "Average Silhouette Width =",
      round(sil_avg, 3)
      ),
    x = "Observation",
    y = "Silhouette Width",
    fill = "Cluster"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    plot.title = element_text(
      size = 16,
      face = "plain"
    ),
    plot.subtitle = element_text(
      size = 10,
      colour = "grey30"
    ),
    legend.title = element_text(face = "plain")
  )

ggsave(
  "C:/Users/Alice/Documents/STAT394-GROUP-3/Figures/PAM.png",
  width = 8,
  height = 6,
  dpi = 600
)

pam_sum_tab |>
gt() |>
tab_header(
title = "PAM Clustering Summary",
subtitle = "Average silhouette width and cluster sizes"
) |>
fmt_number(
columns = Avg_Silhouette,
decimals = 3
) 

top_pam <- pam_sum_tab |>
arrange(desc(Avg_Silhouette)) |>
slice(1:10)

knitr::kable(
top_pam,
caption = "Top ten cluster choices ranked by average silhouette width."
)

```

## H.cluster

```{r manhatten cluster, echo=FALSE}
#| fig-format: png
#| dpi: 300
#| fig-width: 12
#| fig-height: 7

d <- dist(scaled, method = "manhattan")
HiMan <- hclust(d, method = "average")

CC <- cor(d, cophenetic(HiMan))

dend <- as.dendrogram(HiMan)

dend_col <- color_branches(
  dend,
  k = 3,
  col = c("#648FFF", "#DC267F", "#FFB000")
)

dend_data <- dendro_data(dend_col)

fviz_dend(
  HiMan,
  k = 3,
  rect = TRUE,
  rect_border = c("#648FFF", "#DC267F", "#FFB000"),
  k_colors = c("#648FFF", "#DC267F", "#FFB000"),
  color_labels_by_k = FALSE
) +
  labs(
    title = "Hierarchichal Cluster Dendrogram",
    subtitle = paste(
      "Average Linkage | Manhattan Distance | CCC =",
      round(CC, 3)
    ),
    colour = "Cluster"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(colour = "grey30")
  ) 


#saving locally to move to repository

#ggsave(
  #file.path("LOCAL FILE PATH"),
  #width = 10,
  #height = 6,
  #dpi = 300
#)

```

### UMAP

```{r UMAP}
## Trying stuff - Note Gaia couldn't get this to work ALICE

k <- umap(scaled)
PamMan <- pam(scaled, k = 3, metric = "manhattan")

UMAP_plot <- data.frame(
  UMAP1 = k$layout[,1],
  UMAP2 = k$layout[,2],
  Cluster = factor(PamMan$cluster)
  )

UMAP_plot$Sample <- rownames(scaled)

ggplot(UMAP_plot,
       aes(
         UMAP1,
         UMAP2,
         colour = Cluster,
         fill = Cluster
         )
       ) + 
  stat_ellipse(
    geom = "polygon",
    alpha = 0.50,
    colour = NA,
    level = 0.95
    ) +
  geom_point(
    size = 5,
    alpha = 0.8
    ) + 
  geom_text(
    aes(label = Sample),
    colour = "black",
    size = 2.5,
    fface = "bold"
  ) +
  scale_colour_manual(
    values = c(
      "1" = "#648FFF",
      "2" = "#DC267F",
      "3" = "#FFB000"
      ) 
    ) +
  scale_fill_manual(
    values = c(
      "1" = "grey60",
      "2" = "grey60",
      "3" = "grey60"
      )
    ) +
  labs(
    title = "UMAP Projection",
    subtitle = "Manhattan Distance | PAM clustering | k = 3",
    x = "UMAP1",
    y = "UMAP2"
    ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
      plot.title = element_text(
        size = 16,
        face = "plain"
      ),
      plot.subtitle = element_text(
        size = 10,
        colour = "grey30"
      ),
      legend.title = element_text(face = "plain")
    )

ggsave(
  "C:/Users/Alice/Documents/STAT394-GROUP-3/Figures/UMAP.png",
  width = 10,
  height = 8,
  dpi = 600
)


UMAP <- umap(scaled, metric = "manhattan")

#head(UMAP)

plotPamMan <- data.frame(
  UMAP1 = UMAP$layout[,1],
  UMAP2 = UMAP$layout[,2],
  Cluster = factor(PamMan$cluster)
  )

ggplot(plotPamMan, 
       aes(UMAP1, UMAP2, color = cluster, size = medoid)
       ) +
  geom_point(size = 5) +
  theme_classic() +
  labs(title = "UMAP Projection",
       subtitle = "Manhattan Distance | PAM clustering | k = 3"
       )
          
               
```

## Fuzzy Analysis

```{r fuzzy, echo=FALSE}
#| fig-width: 10
#| fig-height: 6

#k-medoids for Fuzzy

Fuzzy_k <- FKM.med(scaled, k = 3, m = 2)

#silhouettes

#Table of Fuzzy_K$U:
knitr::kable(Fuzzy_k$U,
             caption = "Fuzzy K Silhouettes Clusterings",
             format = "pipe")



Fuzzy_Sil <- SIL.F(scaled, Fuzzy_k$U)

Fuzzy_Clust <- apply(Fuzzy_k$U, 1, which.max)

d <- dist(scaled)

sil_each <- silhouette(Fuzzy_Clust, d)

head(sil_each)

plot(sil_each,
     col = my_col[sort(unique(Fuzzy_Clust))],
     border = NA,
     main = "Silhouette Plot")


sil_df <- data.frame(
  obs = 1:nrow(sil_each), 
  width = sil_each[, "sil_width"],
  cluster = factor(sil_each[, "cluster"])
  )

ggplot(sil_df,
       aes(x = reorder(obs, width),
           y = width,
           fill = cluster)) + 
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = mean(sil_df$width),
             linetype = "dashed",
             colour = "red") +
  scale_fill_manual(values = c(
    "#648FFF", "#DC267F", "#FFB000")) +
  labs(title = "Silhouette Plot",
       x = "Observation",
       y = "Silhouette Width") +
  theme_minimal()

plot(sil_each[, "sil_width"],
     pch = 15,
     col = my_col[Fuzzy_Clust],
     xlab = "Observation",
     ylab = "Silhouette Width",
     main = "Per-Observation Silhouettes")
abline(h = 0, lty = 2)



ggplot(LDA_K,
       aes(k, avg_silhouette)) +
  geom_line() +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "Average Silhouette Width by Number of Clusters",
    x = "Number of Clusters",
    y = "Average Silhouette Width"
    )


certainty <- apply(Fuzzy_k$U, 1, max)

ggplot(
  data.frame(
    Obs = 1:length(certainty),
    Certainty = certainty,
    Cluster = factor(Fuzzy_Clust)
  ),
  aes(Obs, Certainty, colour = Cluster)
  ) +
  geom_hline(
    yintercept = 1/3,
    linetype = "dashed",
    colour = "red"
  ) +
  scale_colour_manual( values = c("#648FFF", "#DC267F", "#FFB000")
  )+
  labs(
    caption = "Red dashed line indicates complete ambiguity (1/k)"
    ) +
  geom_point(size = 6) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Fuzzy Membership Certainty",
    subtitle = "PAM Clustering | Manhattan Distance | k = 3",
    x = "Observations",
    y = "Maximum Membership"
    ) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(
      size = 16,
      face = "plain"
    ),
    plot.subtitle = element_text(
      size = 10,
      colour = "grey30"
    ),
    legend.title = element_text(face = "plain")
  )

ggsave(
"C:/Users/Alice/Documents/STAT394-GROUP-3/Figures/Fuzzy.png",
width = 10,
height = 8,
dpi = 600
)

```

### Comparison tables

```{r}
#creating tables to visually compare results from PAM, H cluster, UMAP, Fuzzy.

# PAM

pam_fit <- pam(scaled, k = 3)
pam_size <- table(pam_fit$clustering)
pam_avg_sil <- pam_fit$silinfo$avg.width
pam_clust <- pam_fit$clustering

# Hierarchical

h_clust <- cutree(HiMan, k = 3)
hclust_size <- table(h_clust)
hclust_sil <- silhouette(h_clust, d)
hclust_avg_sil <- mean(hclust_sil[, "sil_width"])

#UMAP clustering
PamMan

umap_size <- table(PamMan$clustering)
umap_avg_sil <- PamMan$silinfo$avg.width
umap_clust <- PamMan$clustering

# Fuzzy
Fuzzy_Clust

fuzzy_size <- table(Fuzzy_Clust)
fuzzy_avg_sil <- mean(
  silhouette(Fuzzy_Clust, d)[, "sil_width"]
  )
fuzzy_clust <- Fuzzy_Clust

# Silhouettes
#pam_sil
#h_sil
#Fuzzy_Sil

# CCC
CC


#Building table

comp_table <- tibble::tibble(
  Method = c(
    "PAM",
    "Hierarchical Clustering",
    "UMAP + PAM",
    "Fuzzy Analysis"
    ),
  Clusters = c(
    3,3,3,3
  ),
  Mean_Size = c(
    mean(pam_size),
    mean(hclust_size),
    mean(umap_size),
    mean(fuzzy_size)
  ),
  Max_size = c(
    max(pam_size),
    max(hclust_size),
    max(umap_size),
    max(fuzzy_size)
  ),
  Min_size = c(
    min(pam_size),
    min(hclust_size),
    min(umap_size),
    min(fuzzy_size)  
  ),
  Avg.Silhouette = c(
    pam_avg_sil,
    hclust_avg_sil,
    umap_avg_sil,
    fuzzy_avg_sil
  ),
  CCC = c(
    NA,
    CC,
    NA,
    NA
  )
)

comp_table |>
  knitr::kable(
    digits = 3,
    caption = "Comparison of Clustering Methods"
    ) 

cluster_compare <- data.frame(
  Sample = rownames(scaled),
  PAM = pam_clust,
  Hierarchical = h_clust,
  UMAP = umap_clust,
  Fuzzy = fuzzy_clust
  )

cluster_compare$Agreement <- apply(
cluster_compare[, 2:5],
1,
function(x) length(unique(x))
)

knitr::kable(
cluster_compare,
caption = "Cluster Assignment by Method"
)

#ARI because ignores arbitrary clustering labels and measures similarity 
#between groupings instead.


adjustedRandIndex(pam_clust, h_clust)

adjustedRandIndex(pam_clust, umap_clust)

adjustedRandIndex(pam_clust, fuzzy_clust)

ari_matrix <- matrix(
  c(
    adjustedRandIndex(pam_clust, pam_clust),
    adjustedRandIndex(pam_clust, h_clust),
    adjustedRandIndex(pam_clust, umap_clust),
    adjustedRandIndex(pam_clust, fuzzy_clust),
    
    adjustedRandIndex(h_clust, pam_clust),
    adjustedRandIndex(h_clust, h_clust),
    adjustedRandIndex(h_clust, umap_clust),
    adjustedRandIndex(h_clust, fuzzy_clust),
    
    adjustedRandIndex(umap_clust, pam_clust),
    adjustedRandIndex(umap_clust, h_clust),
    adjustedRandIndex(umap_clust, umap_clust),
    adjustedRandIndex(umap_clust, fuzzy_clust),
    
    adjustedRandIndex(fuzzy_clust, pam_clust),
    adjustedRandIndex(fuzzy_clust, h_clust),
    adjustedRandIndex(fuzzy_clust, umap_clust),
    adjustedRandIndex(fuzzy_clust, fuzzy_clust)
    ),
  nrow = 4,
  byrow = TRUE
  )

colnames(ari_matrix) <- c("PAM", "H_clust", "UMAP", "Fuzzy")
rownames(ari_matrix) <- c("PAM", "H_clust", "UMAP", "Fuzzy")

round(ari_matrix, 3)

knitr::kable(
round(ari_matrix, 3),
caption = "Adjusted Rand Index Comparison of Clustering Methods"
)


#Creating heatmap


rownames(ari_matrix) <- c("PAM","Hierarchical","UMAP","Fuzzy")
colnames(ari_matrix) <- c("PAM","Hierarchical","UMAP","Fuzzy")

ari_plot <- ari_matrix
diag(ari_plot) <- NA
ari_plot[upper.tri(ari_plot)] <- NA

ari_long <- reshape2::melt(
ari_plot,
na.rm = TRUE
)

ggplot(
  ari_long,
  aes(Var1, Var2, fill = value)
  ) +
  geom_tile(
    colour = "gray80",
    linewidth = 0.3
  ) +
  geom_text(
    aes(
      label = format(round(value, 2), nsmall = 2)),
    colour = "gray20",
    size = 3
    ) +
  scale_fill_gradient2(
    low = "#FE6100",
    mid = "gray90",
    high = "#648FFF",
    midpoint = 0,
    limits = c(-1, 1),
    name = "ARI"
    ) +
  labs(
    title = "Clustering Consensus Matrix",
    subtitle = "Adjusted Rand Index | Clustering Methods",
    x = "",
    y = "",
    fill = "ARI"
    ) +
  coord_fixed() +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
      ),
    plot.title = element_text(
      size = 16,
      face = "plain"
      ),
    plot.subtitle = element_text(
      size = 10,
      colour = "grey30"
      ),
    legend.title = element_text(face = "bold")
    )

ggsave(
  "C:/Users/Alice/Documents/STAT394-GROUP-3/Figures/ARI_heatmap.png",
  width = 8,
  height = 6,
  dpi = 600
  )

```
